---
title: "Davis County Age-Structured Model"
output: rmarkdown::md_document
vignette: >
  %\VignetteIndexEntry{Davis County Age-Structured Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7
)
```

```{r setup}
library(multigroup.vaccine)
library(socialmixr)

library(ggplot2)
```

This vignette demonstrates an age-structured model of transmission within Davis County, Utah. 

First we define the age groups that we want use in our model. We define the `age_limits` vector using the lower limit (minimum age) of each group, starting with 0:

```{r}
# under 1, 1-4, 5-11, 12-13, 14-17, 18-24, 25-44, 45-69, 70 plus
age_limits <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
```

Next we collect census data from Davis County, Utah, using our custom age group choice:

```{r}
davis_data <- getCensusData(
  state_fips = getStateFIPS("Utah"),
  county_name = "Davis County",
  year = 2024,
  age_groups = age_limits,
  csv_path = getCensusDataPath()
)
```

Now we specify immunity levels of our population age groups due to vaccination and prior infection. Let's assume no prior immunity to H5N1.

```{r}
age_immunity <- rep(0, length(age_limits))
```

Now we can begin defining the input arguments to our `finalsize` function so that we can estimate the outbreak size in this county after an introduction from a 13-year-old.

Initial population size of each disease state:

```{r}
popsize <- davis_data$age_pops

initV <- round(age_immunity * popsize)  # initially immune cases 

initI <- rep(0, length(popsize))  # initial infectious cases
initI[4] <- 1                     # assume 1 initial case in the 4th age group (12-13)

initR <- rep(0, length(popsize))  # no recent prior H5 flu survivors
```

Transmission matrix ingredients: contact matrix, relative susceptibility and transmissibility. The contact matrix uses Polymod contact survey data, adjusted for local population distribution:

```{r}
contactmatrix <- contactMatrixPolymod(age_limits, popsize)
relsusc <- rep(1, length(popsize))      # Assume no age differences in susceptibility
reltransm <- rep(1, length(popsize))    # or transmissibility per contact
```

## Basic reproduction number.

H5 flu could have a low level of transmissibility with $R_0<1$ - let's see how outbreak sizes might look for a range of 0.1 to 0.9. The results are from 1,000 simulations and show how many of those simulations resulted in each number of total infections (including the first one).

```{r}
# Define R0 values to inspect
r0_low_vals <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)

# Run simulations
results_low <- do.call(rbind, lapply(r0_low_vals, function(r) {
  fs <- finalsize(popsize, r, contactmatrix, relsusc, reltransm, initR, initI, initV,
                  method = "stochastic", nsims = 1000)
  data.frame(R0 = as.factor(r), Cases = rowSums(fs))
}))

# Visualization: Fixed axis breaks to stop overlapping
ggplot(results_low, aes(x = Cases)) +
  geom_bar(fill = "steelblue", width = 1) + # width=1 removes gaps between discrete integers
  facet_wrap(~R0, labeller = label_both) +
  scale_x_continuous(breaks = c(1, 10, 20, 30), limits = c(0, 35)) + 
  labs(title = "Distribution of Outbreak Sizes (Sub-critical)",
       subtitle = "Simulation results for R0 < 1",
       y = "Frequency",
       x = "Total Cases") +
  theme_minimal()
```
If $R_0 > 1$, there could be a very large outbreak. We can provide a quick estimate of outbreak sizes using our hybrid model:

```{r}
# Define R0 values to inspect
r0_high_vals <- c(1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9)

# Run simulations
results_high <- do.call(rbind, lapply(r0_high_vals, function(r) {
  fs <- finalsize(popsize, r, contactmatrix, relsusc, reltransm, initR, initI, initV,
                  method = "hybrid", nsims = 1000)
  data.frame(R0 = as.factor(r), Cases = rowSums(fs))
}))

# Visualization: 3x3 Grid + Log Scale Bins
ggplot(results_high, aes(x = Cases)) +
  # Manually define log-spaced breaks to ensure the histogram renders cleanly
  geom_histogram(breaks = 10^seq(0, 6, by = 0.25), fill = "firebrick") +
  facet_wrap(~R0, labeller = label_both, ncol = 3) + # Grid layout instead of vertical
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  labs(title = "Bimodal Outcome: Extinction vs. Epidemic",
       subtitle = "R0 > 1 outcomes split into tiny outbreaks or major epidemics",
       y = "Frequency",
       x = "Total Cases (Log Scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Other Diagrams

```{r other-diagrams, warning=FALSE}
# Calculate the probability of over 100 cases across a range of R0 values
r0_range <- seq(0.1, 2.5, by = 0.1)
risk_data <- do.call(rbind, lapply(r0_range, function(r) {
  # Run 500 simulations per R0
  fs <- finalsize(popsize, r, contactmatrix, relsusc, reltransm, initR, initI, initV,
                  method = "hybrid", nsims = 500)

  # Calculate proportion of runs that have over 100 cases
  prob_epidemic <- mean(rowSums(fs) > 100)

  data.frame(R0 = r, Probability = prob_epidemic)
}))

# Plot the Sigmoid Risk Curve
ggplot(risk_data, aes(x = R0, y = Probability)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(size = 3, alpha = 0.6) +
  # Add a reference line where risk begins to skyrocket
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "gray50") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Epidemic Risk Curve",
       subtitle = "Probability of over 100 cases as R0 increases",
       y = "Probability of Epidemic",
       x = "Basic Reproduction Number (R0)") +
  theme_minimal()

```