---
title: "Age distribution effect on $R_0$"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age distribution effect on R0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7
)
```

```{css, echo=FALSE}
body {
  max-width: 1100px;
  margin: auto;
  padding: 1em;
}

table {
  width: 100%;
}
```
This vignette shows the effects of the age distribution of the Short Creek community on
the basic reproduction number $R_0$, compared to other populations with different age distributions.

```{r setup}
library(multigroup.vaccine)
library(socialmixr)
```

## Age groups

```{r age-setup}
agelims <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
```

## Getting city population data for Short Creek community

```{r get-city-data}
# Load city data files
hildale_path <- system.file("extdata", "hildale_ut_2023.csv", package = "multigroup.vaccine")
colorado_city_path <- system.file("extdata", "colorado_city_az_2023.csv", package = "multigroup.vaccine")
centennial_park_path <- system.file("extdata", "centennial_park_az_2023.csv", package = "multigroup.vaccine")

hildale <- getCityData(
  city_name = "Hildale city, Utah",
  csv_path = hildale_path,
  age_groups = agelims
)

colorado_city <- getCityData(
  city_name = "Colorado City town, Arizona",
  csv_path = colorado_city_path,
  age_groups = agelims
)

centennial_park <- getCityData(
  city_name = "Centennial Park CDP, Arizona",
  csv_path = centennial_park_path,
  age_groups = agelims
)

agepops <- round(hildale$age_pops + colorado_city$age_pops + centennial_park$age_pops)
```
## Getting data from Washington County, Utah (where Hildale is located)
```{r get-county-data}
utah_fips <- getStateFIPS("Utah")
census_csv <- getCensusDataPath()
county <- "Washington County"

county_data <- getCensusData(
  state_fips = utah_fips,
  county_name = county,
  year = 2024,
  age_groups = agelims,
  csv_path = census_csv
)

county_pops <- county_data$age_pops
```
## Getting Utah State and national data:

```{r state-and-national}
state_ageprop <- c(0.0143, 0.0572, 0.1119, 0.0330, 0.0658, 0.1164, 0.2844, 0.2360, 0.0840)
state_agepops <- 3503613 * state_ageprop / sum(state_ageprop)

us_ageprop <- c(0.0133, 0.0517, 0.0825, 0.0335, 0.0670, 0.0900, 0.270, 0.304, 0.086)
us_agepops <- 340100000 * us_ageprop / sum(us_ageprop)
```
## School data

```{r enter-school-data}
schoolpops <- c(250, 350, 190, 86, 150, 84, 114, 108, 205)
schoolagegroups <- c(3, 3, 3, 4, 4, 4, 5, 5, 5)
```

## Create contact matrix for Short Creek

```{r contact-matrix}
#Readjust the school populations to match the age data:
for(a in unique(schoolagegroups)){
  inds <- which(schoolagegroups == a)
  schoolpops[inds] <- round(agepops[a] * schoolpops[inds] / sum(schoolpops[inds]))
}
cm <- contactMatrixAgeSchool(agelims, agepops, schoolagegroups, schoolpops, schportion = 0.7)
grouppops <- c(agepops[1:(min(schoolagegroups)-1)],
               schoolpops,
               agepops[(max(schoolagegroups)+1):length(agepops)])
```

## Create contact matrices for other populations

```{r other-contact}
mijpolymod <- contactMatrixPolymod(agelims)
mijwashingtoncounty <- contactMatrixPolymod(agelims, county_pops)
mijutahstate <- contactMatrixPolymod(agelims, state_agepops)
mijusa <- contactMatrixPolymod(agelims, us_agepops)
```

## Calculate $R_0$ adjustment factors

```{r R0-factors}
eigpolymod <- eigen(mijpolymod)$values[1]
R0factorShortCreek <- eigen(cm)$values[1] / eigpolymod
R0factorWashingtonCounty <- eigen(mijwashingtoncounty)$values[1] / eigpolymod
R0factorUtahState <- eigen(mijutahstate)$values[1] / eigpolymod
R0factorUSA <- eigen(mijusa)$values[1] / eigpolymod

R0vals <- 6:12

R0local <- R0vals * R0factorShortCreek
R0county <- R0vals * R0factorWashingtonCounty
R0state <- R0vals * R0factorUtahState
R0national <- R0vals * R0factorUSA

knitr::kable(data.frame(R0Polymod = R0vals,
                        R0USA = round(R0national, 1),
                        R0Utah = round(R0state, 1),
                        R0County = round(R0county, 1),
                        R0ShortCreek = round(R0local, 1)),
             row.names = FALSE, format = "markdown")
```
