---
title: "Using Census Data Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Census Data Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

```{css, echo=FALSE}
body {
  max-width: 1100px;
  margin: auto;
  padding: 1em;
}

table {
  width: 100%;
}
```

## Introduction

This vignette demonstrates how to use the census data functions in the `multigroup.vaccine` package. These functions allow you to:

1. Look up state FIPS codes
2. List counties within a state
3. Retrieve population data by age groups
4. Retrieve city-level population data with flexible age grouping
5. Run outbreak models with easily accessible real-world data

We'll walk through a complete example using Utah county data.

```{r setup}
library(multigroup.vaccine)
library(socialmixr)

# Get the path to the included census data file
census_csv <- getCensusDataPath()
```

## Step 1: Getting State FIPS Codes

The `getStateFIPS()` function looks up the FIPS code for a state by name. FIPS codes are standardized identifiers used by the U.S. Census Bureau.

```{r state-fips}
# Get FIPS code for Utah
utah_fips <- getStateFIPS("Utah")
cat("Utah FIPS code:", utah_fips, "\n")

# You can also try other states
california_fips <- getStateFIPS("California")
cat("California FIPS code:", california_fips, "\n")

texas_fips <- getStateFIPS("Texas")
cat("Texas FIPS code:", texas_fips, "\n")
```

## Step 2: Listing Counties

The `listCounties()` function returns all counties in a state. This is useful for exploring available data.

```{r list-counties}
# List all counties in Utah
utah_counties <- listCounties(
  state_fips = utah_fips,
  csv_path = census_csv
)

cat("Counties in Utah:\n")
print(utah_counties)

cat("\nTotal number of counties:", length(utah_counties), "\n")
```

## Step 3: Retrieving Population Data

The `getCensusData()` function retrieves population data for a specific county, organized by age groups.

```{r get-census-data}
# Define age groups for analysis
# These represent: 0-4, 5-11, 12-17, 18-24, 25-44, 45-64, 65+
age_limits <- c(0, 5, 12, 18, 25, 45, 65)

# Get data for Washington County, Utah
washington_data <- getCensusData(
  state_fips = utah_fips,
  county_name = "Washington County",
  year = 2024,
  age_groups = age_limits,
  csv_path = census_csv
)

# Display the results
cat("County:", washington_data$county, "\n")
cat("Year:", washington_data$year, "\n")
cat("Total population:", format(washington_data$total_pop, big.mark = ","), "\n\n")

cat("Age distribution:\n")
for (i in seq_along(washington_data$age_labels)) {
  pct <- 100 * washington_data$age_pops[i] / washington_data$total_pop
  cat(sprintf("  %s: %s (%.1f%%)\n",
              washington_data$age_labels[i],
              format(washington_data$age_pops[i], big.mark = ","),
              pct))
}
```

## Step 4: Visualizing Age Distribution

Let's visualize the age distribution to better understand the population structure.

```{r plot-age-distribution, fig.alt="Bar chart showing age distribution percentages for Washington County, Utah. Each bar represents an age group with percentage of total population labeled above."}
# Create a bar plot of age distribution
age_percentages <- 100 * washington_data$age_pops / washington_data$total_pop

barplot(age_percentages,
        names.arg = washington_data$age_labels,
        main = paste("Age Distribution -", washington_data$county),
        xlab = "Age Group",
        ylab = "Percentage of Population",
        col = "steelblue",
        las = 2,
        ylim = c(0, max(age_percentages) * 1.1))

# Add percentage labels on top of bars
text(x = seq_along(age_percentages) * 1.2 - 0.5,
     y = age_percentages + 1,
     labels = sprintf("%.1f%%", age_percentages),
     pos = 3,
     cex = 0.8)
```

## Step 5: Comparing Multiple Counties

Let's compare the age distributions of three different Utah counties.

```{r compare-counties, fig.alt="Side-by-side bar chart comparing age distribution percentages across Salt Lake County (coral), Utah County (steel blue), and Washington County (light green). Each age group shows three bars representing the three counties."}
# Get data for three counties
counties_to_compare <- c("Salt Lake County", "Utah County", "Washington County")
county_data_list <- list()

for (county in counties_to_compare) {
  county_data_list[[county]] <- getCensusData(
    state_fips = utah_fips,
    county_name = county,
    year = 2024,
    age_groups = age_limits,
    csv_path = census_csv
  )
}

# Create comparison matrix
comparison_matrix <- matrix(0, nrow = length(counties_to_compare), 
                            ncol = length(age_limits))
colnames(comparison_matrix) <- washington_data$age_labels
rownames(comparison_matrix) <- c("Salt Lake", "Utah", "Washington")

for (i in seq_along(counties_to_compare)) {
  county_name <- counties_to_compare[i]
  data <- county_data_list[[county_name]]
  comparison_matrix[i, ] <- 100 * data$age_pops / data$total_pop
}

# Plot comparison
barplot(comparison_matrix,
        beside = TRUE,
        main = "Age Distribution Comparison Across Utah Counties",
        xlab = "Age Group",
        ylab = "Percentage of Population",
        col = c("coral", "steelblue", "lightgreen"),
        legend.text = rownames(comparison_matrix),
        args.legend = list(x = "topright", bty = "n"),
        las = 2)
```

## Step 6: Using City-Level Data (getCityData)

In addition to census data, the package includes functions to work with city-level population data from ACS (American Community Survey) estimates. Let's demonstrate using Hildale, UT as an example.

### Example 1: Default 5-year Age Groups

```{r city-data-5year}
# Get path to Hildale data
hildale_path <- system.file("extdata", "hildale_ut_2023.csv", package = "multigroup.vaccine")

# Load with default 5-year age groups (0-4, 5-9, 10-14, ...)
hildale_5yr <- getCityData(
  city_name = "Hildale city, Utah",
  csv_path = hildale_path
)

cat("Hildale, UT - 5-year Age Groups\n")
cat("================================\n")
cat("Total population:", format(hildale_5yr$total_pop, big.mark = ","), "\n\n")

cat("Age distribution:\n")
for (i in seq_along(hildale_5yr$age_labels)) {
  pct <- 100 * hildale_5yr$age_pops[i] / hildale_5yr$total_pop
  cat(sprintf("  %s: %s (%.1f%%)\n",
              hildale_5yr$age_labels[i],
              format(hildale_5yr$age_pops[i], big.mark = ","),
              pct))
}
```

### Example 2: Custom Age Groups for School-Based Analysis

Now let's use custom age groups that align with school levels: pre-school (0-4), elementary (5-11), middle school (12-13), high school (14-17), and adult groups.

```{r city-data-custom}
# Define school-aligned age groups
school_age_groups <- c(0, 5, 12, 14, 18, 25, 45, 65)

hildale_school <- getCityData(
  city_name = "Hildale city, Utah",
  csv_path = hildale_path,
  age_groups = school_age_groups
)

cat("\nHildale, UT - School-Aligned Age Groups\n")
cat("========================================\n")
cat("Total population:", format(hildale_school$total_pop, big.mark = ","), "\n\n")

cat("Age distribution:\n")
for (i in seq_along(hildale_school$age_labels)) {
  pct <- 100 * hildale_school$age_pops[i] / hildale_school$total_pop
  cat(sprintf("  %s: %s (%.1f%%)\n",
              hildale_school$age_labels[i],
              format(hildale_school$age_pops[i], big.mark = ","),
              pct))
}
```

### Visualizing the Comparison

```{r compare-aggregations, fig.alt="Two side-by-side bar charts comparing age grouping methods for Hildale, UT. Left panel shows 5-year age groups in steel blue. Right panel shows school-aligned age groups in coral. Both display percentage of population on y-axis."}
# Create a comparison visualization
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# Plot 5-year groups
age_pct_5yr <- 100 * hildale_5yr$age_pops / hildale_5yr$total_pop
barplot(age_pct_5yr,
        names.arg = hildale_5yr$age_labels,
        main = "5-Year Age Groups",
        xlab = "Age Group",
        ylab = "% of Population",
        col = "steelblue",
        las = 2,
        cex.names = 0.7,
        ylim = c(0, max(age_pct_5yr) * 1.2))

# Plot school-aligned groups
age_pct_school <- 100 * hildale_school$age_pops / hildale_school$total_pop
barplot(age_pct_school,
        names.arg = hildale_school$age_labels,
        main = "School-Aligned Age Groups",
        xlab = "Age Group",
        ylab = "% of Population",
        col = "coral",
        las = 2,
        cex.names = 0.7,
        ylim = c(0, max(age_pct_school) * 1.2))

par(mfrow = c(1, 1))
```

This demonstrates how `getCityData()` provides flexibility in age grouping. The 5-year groups follow standard ACS categories, while custom groups can be defined to match specific analysis needs (e.g., school grades, epidemiological contact patterns).

## Summary

This vignette demonstrated the key population data functions in the `multigroup.vaccine` package:

1. **`getStateFIPS()`**: Look up state FIPS codes
2. **`listCounties()`**: List all counties in a state
3. **`getCensusData()`**: Retrieve county-level population data by age groups from U.S. Census
4. **`getCityData()`**: Retrieve city-level population data from ACS with flexible age grouping

These functions make it easy to work with real population data from both census (county-level) and ACS (city-level) sources. The flexible age grouping options allow you to structure data according to your specific analysis needs, whether for epidemic modeling, demographic analysis, or other applications. You can apply these same techniques to any state and county in the United States.
