---
title: "Run Model in the console"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Model in the console}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multigroup.vaccine)
```

Consider a population with the following parameters:

-   Two sub-populations, one with size 80000 (A) and the other with size 20000 (B)

-   At time 0, vaccinate 30% of A and 20% of B

-   Set the reproductive number 1.75

-   Recovery rate of 1/7 creates a time to symptom resolution of 1 week.

-   Contact ratio is 1.7. This is the overall contact rate for B over the overall contact rate for A. This means that an individual in Group B makes 1.7 contacts for every 1 made by an individual in group A.

-   Contact within group is set to 0.2 for A, 0.5 for B. This is the fraction of all contacts that are A-\>A, or B-\>B respectively.

-   Susceptibility ratio is the susceptibility of individuals in population B over that of individuals in population A. With susceptibility = 1, they are equally susceptible.

```{r}
multigroup.vaccine::getFinalSize(vacTime = 0,
  vacPortion = c(0.3, 0.2),
  popSize = c(80000, 20000),
  R0 = 1.75,
  recoveryRate = 1 / 7,
  contactRatio = 1.7,
  contactWithinGroup = c(0.2, 0.5),
  suscRatio = 1)
```

```
[1] 15145.987  8710.351
```

More than half as many individuals in population B are infected at the end of the simulation, despite the fact that they are outnumbered 4 to 1 by population A.
