---
title: "Sensitivity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multigroup.vaccine)
```

Consider a population with the following parameters:

-   Two sub-populations sized 80000 (A) and 20000 (B)

-   At time zero 20% of both A and B are vaccinated

-   R-Naught is 1.75, infected individuals take a week to recover.

-   With a contact ratio of 2, Group B makes twice as many contacts as group A.

-   75% of Group A's contacts are within Group A. 50% of Group B's contacts are within Group B

-   Susceptibility ratio varies from 0.5 to 2. At 0.5 Group A is twice as susceptible to infection as Group B. At 2, Group A is half as susceptible to infection as Group A.

```{r}
index <- vector()
groupA <- vector()
groupB <- vector()
popA <- 80000
popB <- 20000

# For each
for (i in seq(0.5, 2, 0.01)) {

  ret <- getFinalSize(vacTime = 0,
    vacPortion = c(0.2, 0.2),
    popSize = c(popA, popB),
    R0 = 1.75,
    recoveryRate = 1 / 7,
    contactRatio = 2,
    contactWithinGroup = c(0.75, 0.5),
    suscRatio = i)

  if (ret[1] > 1000) {
    index <- c(index, i)
    groupA <- c(groupA, ret[1] / popA)
    groupB <- c(groupB, ret[2] / popB)
  }


}

print(max(max(groupA), max(groupB)))
plot(index, groupA, type = "l", xlab = "Group B / Group A Susceptibility Ratio", ylab = "Infection Prevalence", ylim = c(0, max(max(groupA), max(groupB))))
lines(index, groupB, type = "l", col = "red")
legend("topright", legend = c("Group A Prevalence", "Group B Prevalence"), col = c("black", "red"), lty = 1)
```
