---
title: "Measles Outbreak Simulations in Utah Counties"
date: "May 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measles Outbreak Simulations in Utah Counties}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in any specific location.
DATA: The vaccination data (county level) is not public and so is not included in the package.

Executive Summary

This vignette provides code for setting up and running stochastic measles outbreak simulations for
every county in Utah. The simulation inputs include demographic mixing patterns and local vaccination coverage assumptions based on school vaccination data.

## Data Initialization & Demographic Calibration

This section processes county-level census data, school vaccination records, and age-structured contact matrices (Polymod). The baseline pathogen transmissibility (R0​) is calibrated to account for the unique age distribution of each county.

```{r setup-libs, cache=FALSE}
library(multigroup.vaccine)
library(socialmixr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)
library(scales)

# Load raw data
county_vax_data <- read.csv(file.path("../../data-raw/county_vax_data.csv"))
unique_counties <- unique(county_vax_data$county)

# Infer age structures based on population ratios
age_groups_by_county <- list()
for (county_name in unique_counties) {
  county_data <- county_vax_data[county_vax_data$county == county_name, ]
  el_pop <- county_data$pop[county_data$level == 1]
  middle_pop <- county_data$pop[county_data$level == 2]
  high_pop <- county_data$pop[county_data$level == 3]
  
  if (length(middle_pop) > 0 && length(high_pop) > 0 && high_pop > 0) {
    middle_to_high_ratio <- middle_pop / high_pop
    el_to_middle_ratio <- if(length(el_pop) > 0 && middle_pop > 0) el_pop / middle_pop else NA

    if (middle_to_high_ratio >= 0.8 && middle_to_high_ratio <= 1.2) {
      if (!is.na(el_to_middle_ratio) && el_to_middle_ratio < 1.3) {
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 10, 14, 18, 25, 45, 70)
      } else {
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 15, 18, 25, 45, 70)
      }
    } else if (middle_to_high_ratio <= 0.6) {
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
    } else {
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
    }
  } else {
    age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
  }
}
age_groups_by_county$Online <- NULL

# Fetch Census Data
utah_counties <- getAllCountiesData(
  state_fips = getStateFIPS("Utah"),
  year = 2024,
  age_groups_by_county = age_groups_by_county,
  csv_path = getCensusDataPath(),
  verbose = FALSE
)

# Calculate Immunity Levels & Calibrate R0
R0_baseline <- 10
R0_baseline_range <- seq(6, 12, by=1)
reference_matrix <- contactMatrixPolymod(c(0, 1, 5, 11, 14, 18, 25, 45, 70))
reference_eigenvalue <- eigen(reference_matrix)$values[1]

county_immunity <- list()
county_contact_matrices <- list()
county_R0_factors <- data.frame(county = character(), R0_factor = numeric(), R0_effective = numeric())

for (county_name in names(utah_counties)) {
  county_name_short <- gsub(" County", "", county_name)
  county_data <- utah_counties[[county_name]]
  vax_data <- county_vax_data[county_vax_data$county == county_name_short, ]
  
  el_vax_rate <- vax_data$vax[vax_data$level == 1] / vax_data$pop[vax_data$level == 1]
  middle_vax_rate <- vax_data$vax[vax_data$level == 2] / vax_data$pop[vax_data$level == 2]
  high_vax_rate <- vax_data$vax[vax_data$level == 3] / vax_data$pop[vax_data$level == 3]
  pre_k_vax_rate <- el_vax_rate * 0.9  
  
  pre_k_immunity <- pre_k_vax_rate * 0.93
  el_immunity <- el_vax_rate * 0.97
  middle_immunity <- middle_vax_rate * 0.97
  high_immunity <- high_vax_rate * 0.97

  interpolate_immunity <- function(age, start_age, end_age, start_imm, end_imm) {
    slope <- (end_imm - start_imm) / (end_age - start_age)
    return(start_imm + slope * (age - start_age))
  }
  
  imm_18_25 <- interpolate_immunity(21.5, 18, 70, high_immunity, 1.0)
  imm_25_45 <- interpolate_immunity(35, 18, 70, high_immunity, 1.0)
  imm_45_70 <- interpolate_immunity(57.5, 18, 70, high_immunity, 1.0)

  county_immunity[[county_name]] <- c(0, pre_k_immunity, el_immunity, middle_immunity, high_immunity, imm_18_25, imm_25_45, imm_45_70, 1.0)
  
  # Contact Matrix & Eigenvalue adjustment
  county_matrix <- contactMatrixPolymod(age_groups_by_county[[county_name_short]], county_data$age_pops)
  county_contact_matrices[[county_name_short]] <- county_matrix
  
  R0_factor <- Re(eigen(county_matrix)$values[1]) / Re(reference_eigenvalue)
  county_R0_factors <- rbind(county_R0_factors, data.frame(
    county = county_name_short, R0_factor = R0_factor, R0_effective = R0_baseline * R0_factor
  ))
}
```
2. Stochastic Simulation

We run 1000 independent simulations per county (assuming introduction by a young adult 18-25), for
each assumed value of the reference R0.

```{r stochastic-sims, cache=TRUE}
nsims_stochastic <- 1000

results_list <- vector("list", length(R0_baseline_range) * length(utah_counties))
sim_list     <- vector("list", length(R0_baseline_range) * length(utah_counties))
idx <- 1L

simfun <- function(R0, county_name, nsims, nthreads, ...){
  c_short <- gsub(" County", "", county_name)
  popsize <- utah_counties[[county_name]]$age_pops
  initV <- round(county_immunity[[county_name]] * popsize)
  initI <- rep(0, length(popsize)); initI[6] <- 1  # Intro in young adults
  R0_eff <- R0 * county_R0_factors$R0_factor[county_R0_factors$county == c_short]
  
  # Run pure stochastic simulation
  finalsize(popsize, R0_eff, county_contact_matrices[[c_short]],
            rep(1, length(popsize)), rep(1, length(popsize)),
            rep(0, length(popsize)), initI, initV,
            method = "stochastic", nsims = nsims, nthreads = nthreads, ...)
}

for (R0_stochastic in R0_baseline_range) {
  for (county_name in names(utah_counties)) {
    c_short <- gsub(" County", "", county_name)
    
    # Run pure stochastic simulation
    fs_stochastic <- simfun(R0_stochastic, county_name, nsims_stochastic, nthreads = 1)
    total_infected <- rowSums(fs_stochastic)
    
    # Store all sims
    sim_list[[idx]] <- data.frame(
      baseline_R0 = R0_stochastic,
      county = c_short,
      total_cases = total_infected
    )
    idx <- idx + 1L
  }
}

sim_long <- bind_rows(sim_list)
```

## Full Outbreak Distribution

All 1000 simulations per county are shown. A bimodal distribution — a spike closer to zero (chains that fizzle out) and a heavier right tail (sustained outbreaks) — is a hallmark of stochastic outbreak size distribution for populations above the herd immunity threshold.

```{r}
if(nrow(sim_long) > 0) {
  # Order by median outbreak size at R0 = 10
  order_all <- sim_long %>% filter(baseline_R0 == 10) %>% group_by(county) %>% 
    summarise(med = median(total_cases)) %>% arrange(med) %>% pull(county)
  all_sim_counties <- unique(sim_long$county)
  order_all_full <- c(order_all, sort(setdiff(all_sim_counties, order_all)))

  p_dots <- plot_ly()
  for (r0_val in R0_baseline_range) {
    df_sub <- sim_long %>%
      filter(baseline_R0 == r0_val)

    # Count exact matches within county
    df_sub <- df_sub %>%
    group_by(county, total_cases) %>%
    mutate(n_equal = n()) %>%
    ungroup()

    # Count greater-than values within county
    df_sub$n_greater <- sapply(
      seq_len(nrow(df_sub)),
      function(i) {
        sum(df_sub$county == df_sub$county[i] & df_sub$total_cases > df_sub$total_cases[i])
      })
    df_sub$county_f <- factor(df_sub$county, levels = order_all_full)

    p_dots <- p_dots %>% add_trace(data = df_sub, x = ~total_cases, y = ~county_f,
                                   type = "scatter", mode = "markers",
                                   marker = list(color = "#000000", size = 4, opacity = 0.25),
    hovertext = ~paste0(county,
                        "<br>= ", scales::comma(total_cases), " ",
                        ifelse(total_cases == 1, "case", "cases"),
                        ": ", n_equal, " ",
                        ifelse(n_equal == 1, "simulation", "simulations"),
                        "<br>> ", scales::comma(total_cases), " ",
                        ifelse(total_cases == 1, "case", "cases"),
                        ": ", n_greater, " ",
                        ifelse(n_greater == 1, "simulation", "simulations")
                        ),
    hoverinfo = "text",
    visible = (r0_val == 10),
    name = paste0("R0 = ", r0_val)
    )
  }
  steps_dots <- lapply(seq_along(R0_baseline_range), function(i) {
    vis <- rep(FALSE, length(R0_baseline_range))
    vis[i] <- TRUE
    list(method = "restyle", args = list("visible", as.list(vis)),
         label = as.character(R0_baseline_range[i]))
  })

  p_dots %>% layout(
    title = paste0("Full Distribution of Outbreak Final Size: ",
                   scales::comma(nsims_stochastic),
                   " Simulations"),
    xaxis = list(title = "Total Cases per Simulation (log scale)", type = "log"),
    yaxis = list(title = ""),
    sliders = list(list(active = 2, currentvalue = list(prefix = "Baseline R0 = "),
                        steps = steps_dots)),
    showlegend = FALSE, margin = list(l = 100)
  )
} else {
  print("No simulation results.")
}
```

## Export Simulation Outputs for External Risk Scoring

This vignette now stops at simulation outputs and exports reusable datasets for downstream risk-score development in a separate notebook.

```{r export-simulation-outputs, cache=FALSE}
school_vax_summary <- county_vax_data %>%
  filter(county != "Online") %>%
  group_by(county) %>%
  summarise(
    total_enrolled = sum(pop),
    total_vaccinated = sum(vax),
    school_vax_rate = ifelse(total_enrolled > 0, total_vaccinated / total_enrolled, NA_real_),
    .groups = "drop"
  )

# These county-level metrics are derived from the exact same immunity vectors used
# to initialize the stochastic simulations, so the frontend can display them
# directly without recomputing the simulation setup logic.
county_static_metrics <- bind_rows(lapply(names(utah_counties), function(county_name) {
  county_short <- gsub(" County", "", county_name)
  popsize <- utah_counties[[county_name]]$age_pops
  initV <- round(county_immunity[[county_name]] * popsize)
  total_pop <- sum(popsize)
  immune_population <- sum(initV)
  susceptible_population <- total_pop - immune_population

  data.frame(
    county = county_short,
    population = total_pop,
    immune_population = immune_population,
    susceptible_population = susceptible_population,
    county_immunity_pct = ifelse(total_pop > 0, immune_population / total_pop, NA_real_),
    county_susceptible_pct = ifelse(total_pop > 0, susceptible_population / total_pop, NA_real_),
    stringsAsFactors = FALSE
  )
})) %>%
  left_join(school_vax_summary, by = "county") %>%
  left_join(
    county_R0_factors %>% select(county, R0_factor),
    by = "county"
  )

if (nrow(sim_long) > 0) {
  simulation_stats_by_county <- sim_long %>%
    group_by(baseline_R0, county) %>%
    summarise(
      nsims = n(),
      mean_cases = mean(total_cases),
      median_cases = median(total_cases),
      p90_cases = as.numeric(quantile(total_cases, 0.90)),
      p95_cases = as.numeric(quantile(total_cases, 0.95)),
      max_cases = max(total_cases),
      prob_large_outbreak = mean(is_large_outbreak),
      prob_fizzle = mean(total_cases == 0),
      .groups = "drop"
    )
} else {
  simulation_stats_by_county <- data.frame(
    baseline_R0 = numeric(),
    county = character(),
    nsims = numeric(),
    mean_cases = numeric(),
    median_cases = numeric(),
    p90_cases = numeric(),
    p95_cases = numeric(),
    max_cases = numeric(),
    prob_large_outbreak = numeric(),
    prob_fizzle = numeric()
  )
}

simulation_stats_statewide <- stochastic_results %>%
  group_by(baseline_R0) %>%
  summarise(
    n_counties = n(),
    n_vulnerable = sum(vax_repnum >= 1),
    n_herd_immunity = sum(vax_repnum < 1),
    mean_expected_cases = mean(expected_cases),
    median_expected_cases = median(expected_cases),
    mean_prob_large = mean(prob_large_outbreak),
    .groups = "drop"
  )

export_dir <- file.path("outputs", "utah_county_risk")
if (!dir.exists(export_dir)) dir.create(export_dir, recursive = TRUE)

write.csv(stochastic_results,
          file.path(export_dir, "stochastic_results_summary.csv"),
          row.names = FALSE)
write.csv(sim_long,
          file.path(export_dir, "stochastic_results_all_runs.csv"),
          row.names = FALSE)
write.csv(simulation_stats_by_county,
          file.path(export_dir, "simulation_stats_by_county.csv"),
          row.names = FALSE)
write.csv(simulation_stats_statewide,
          file.path(export_dir, "simulation_stats_statewide.csv"),
          row.names = FALSE)
write.csv(county_static_metrics,
          file.path(export_dir, "county_static_metrics.csv"),
          row.names = FALSE)
write.csv(vaxrepnum_sensitivity,
          file.path(export_dir, "vaxrepnum_sensitivity.csv"),
          row.names = FALSE)
write.csv(threshold_counts,
          file.path(export_dir, "threshold_counts.csv"),
          row.names = FALSE)

save(
  stochastic_results,
  sim_long,
  simulation_stats_by_county,
  simulation_stats_statewide,
  county_static_metrics,
  vaxrepnum_sensitivity,
  threshold_counts,
  county_R0_factors,
  R0_baseline,
  R0_baseline_range,
  large_outbreak_threshold_frac,
  large_outbreak_min_cases,
  nsims_stochastic,
  file = file.path(export_dir, "utah_county_simulation_bundle.RData")
)

cat("Saved simulation outputs to:", normalizePath(export_dir), "\n")
```
