---
title: "Measles Reproduction Number Analysis for Utah Counties"
date: "May 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measles Reproduction Number Analysis for Utah Counties}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in any specific location.
DATA: The vaccination data (county level) is not public and so is not included in the package.

Executive Summary

This vignette provides code for analyzing measles reproduction numbers for every county in Utah. The inputs include demographic mixing patterns and local vaccination coverage assumptions based on school vaccination data.

```{r setup-libs, cache=FALSE}
library(multigroup.vaccine)
library(socialmixr)
```

## Data Initialization

First we read the county vaccination data, which contains the school-level enrollment statistics and vaccination statistics for each county, and the county census data, which contains single-age population statistics.

```{r load-data, cache=FALSE}

# Load raw data
county_vax_data <- read.csv(file.path("../../data-raw/county_data_Utah_2025.csv"))
#unique_counties <- unique(county_vax_data$county)
county_pop_data <- getAllCountiesData(
  state_fips = getStateFIPS("Utah"),
  year = 2024,
  csv_path = getCensusDataPath()
)
```

## School-level age groups by county

Next we convert the school levels (elementary, middle school / junior high school, and high school) from the vaccination data of each county into age levels. Different counties tend have different grades associated with the three levels. This process is discussed in more detail in the "School Age Group Analysis" vignette.

```{r define-age-groups}
grade_groups_by_county <- list()

# Grade - level combinations to test
elem_mid <- cbind(c(4, 4, 5, 5, 6, 6, 7, 7),
                  c(5, 6, 4, 5, 3, 4, 2, 3))
  
for (county in unique(county_vax_data$county)) {

    #get age 5-17 census population data for the county
    schoolagepop <- county_pop_data[[paste0(county, " County")]]$data$TOT_POP[6:18]
    sapfrac <- schoolagepop / sum(schoolagepop)
    
    cvd <- county_vax_data[county_vax_data$county == county, ]
  
    # Get population for each level
    elem_popfrac <- cvd$popfrac[cvd$level == 1]
    mid_popfrac <- cvd$popfrac[cvd$level == 2]
    high_popfrac <- cvd$popfrac[cvd$level == 3]
    
    err <- rep(0, nrow(elem_mid))
    for(i in 1:nrow(elem_mid)){
      ielem <- 1:elem_mid[i,1]
      imid <- (elem_mid[i,1] + 1):rowSums(elem_mid)[i]
      ihigh <- (rowSums(elem_mid)[i] + 1):13
      err[i] <- (sum(schoolagepop[ielem])*(elem_popfrac - sum(sapfrac[ielem])))^2 +
                (sum(schoolagepop[imid])*(mid_popfrac - sum(sapfrac[imid])))^2 +
                (sum(schoolagepop[ihigh])*(high_popfrac - sum(sapfrac[ihigh])))^2
    }
    bestrow <- which.min(err)
    grade_groups_by_county[[county]] <- c(0, elem_mid[bestrow, 1],
                                          rowSums(elem_mid)[bestrow])
}
```


## Age and vaccination / immunity levels by county
Next we define full age groupings and vaccination levels for each county. Age groupings are defined by a vector for the minimum age in years of each group, starting with 0. 

For pre-school groups, we define an under-age-1 group and an age 1-to-4 group, assuming that age-0 individuals have received no MMR doses and age 1-to-4 received a single MMR dose at a proportion assumed for each county.

For school-age groups, we use the `grade_groups_by_county` data from the previous section above, which defines the starting grade levels (0 for kindergarten) of the three levels. We add 5 to these numbers to represent our assumption that age 5 to 17 represents kindergarten through 12th grade. The school vaccination and immunity data are applied directly to these groups, assuming the given rate for two doses of MMR and for prior infection. 

For adult groups, we split into 4-year age groups ranging from an age 18-21 group up to an age 66-69 age group, and finally a 70-and-older age group. For vaccination, we assume a linear interpolation from the high school vaccination percentage of that county up to 100%, where the 100% immunity in the 70+ age group derives from prior infection as children before a measles vaccine was available. 

```{r age-group-data}

age_groups_pre <- c(0, 1)
age_groups_adult <- seq(18, 70, 4)

min_age_adult <- min(age_groups_adult)
age_range_adult <- diff(range(age_groups_adult))
adult_midpoints <-
  (age_groups_adult[-length(age_groups_adult)] + age_groups_adult[-1]) / 2

interpolate_vac <- function(start_vac, end_vac) {
    slope <- (end_vac - start_vac) / age_range_adult
    return(start_vac + slope * (adult_midpoints - min_age_adult))
}

county_df <- data.frame(county = character(), min_age = numeric(), population = numeric(), MMR2 = numeric(), MMR1 = numeric(), priorinf = numeric())

for (county_name in names(county_pop_data)) {
  county_name_short <- gsub(" County", "", county_name)
  county_data <- county_pop_data[[county_name]]
  vax_data <- county_vax_data[county_vax_data$county == county_name_short, ]
  min_age <- c(age_groups_pre,
               5 + grade_groups_by_county[[county_name_short]],
               age_groups_adult)

  el_vax_rate <- vax_data$vaxrate[vax_data$level == 1]
  mid_vax_rate <- vax_data$vaxrate[vax_data$level == 2]
  high_vax_rate <- vax_data$vaxrate[vax_data$level == 3]
  pre_k_vax_rate <- el_vax_rate * 0.9  
  
  el_prior_rate <- vax_data$priorrate[vax_data$level == 1]
  mid_prior_rate <- vax_data$priorrate[vax_data$level == 2]
  high_prior_rate <- vax_data$priorrate[vax_data$level == 3]
  
  adult_vax_rates <- interpolate_vac(high_vax_rate, 1)

  MMR2 <- c(rep(0, length(age_groups_pre)), el_vax_rate, mid_vax_rate, high_vax_rate,
            adult_vax_rates, 0)
  MMR1 <- c(0, rep(pre_k_vax_rate, length(age_groups_pre)-1), rep(0, length(MMR2) - 2))
  prior <- c(rep(0, length(age_groups_pre)), el_prior_rate, mid_prior_rate, high_prior_rate,
             rep(0, length(adult_vax_rates)), 1)
  
  county_df <- rbind(county_df, 
                     data.frame(county = rep(county_name_short, length(MMR2)),
                                min_age = min_age,
                                MMR2 = MMR2, MMR1 = MMR1, priorinf = prior))
}
```
  
##Contact matrix construction

```{r}  
county_contact_matrices <- list()
county_group_pops <- rep(0, nrow(county_df))
ind <- 1
for (county_name in names(county_pop_data)) {
  pop_data <- county_pop_data[[county_name]]
  county_name_short <- gsub(" County", "", county_name)
  df <- county_df[county_df$county == county_name_short, ]
  pop_data_agg <- aggregateByAgeGroups(ages = pop_data$ages,
                                       pops = pop_data$age_pops,
                                       age_groups = df$min_age)
                                         
  county_matrix <- contactMatrixPolymod(df$min_age, pop_data_agg$pops)
  county_contact_matrices[[county_name_short]] <- county_matrix
  county_group_pops[ind:(ind + length(pop_data_agg$pops)-1)] <- pop_data_agg$pops
  ind <- ind + length(pop_data_agg$pops)
}
county_df$population <- county_group_pops
county_total_pop <- aggregate(list(total_pop = county_df$population), list(county = county_df$county), sum)
```
## County vaccination and immunity rankings

```{r county_vax_rank}
school_vax <- 
  aggregate(list(schoolvax = county_vax_data$vaxrate * county_vax_data$popfrac),
            list(county = county_vax_data$county), sum)
school_vax$schoolvax_rank <- rank(school_vax$schoolvax)

imm <- (county_df$MMR2 * 0.97 + county_df$MMR1 * 0.93 + county_df$priorinf) *
  county_df$population

total_pop <- aggregate(list(totalpop = county_df$population),
                       list(county = county_df$county), sum)
pop_immunity <- 
  aggregate(list(popimm = imm),
            list(county = county_df$county), sum)

pop_immunity$popimm <- pop_immunity$popimm / total_pop$totalpop
pop_immunity$popimm_rank <- rank(pop_immunity$popimm)
```
## Local R0 by county

```{r}
county_R0_factors <- data.frame(county = character(), R0_factor = numeric(), R0_effective = numeric())
R0_baseline <- 8
R0_baseline_range <- seq(6, 12, by=1)

grade_groups <- unique(grade_groups_by_county)

reference_eigenvalues <- rep(0, length(grade_groups))
for(i in seq_along(grade_groups)){
  reference_matrix <- contactMatrixPolymod(c(age_groups_pre,
                                             5 + grade_groups[[i]],
                                             age_groups_adult))
  reference_eigenvalues[i] <- eigen(reference_matrix)$values[1]
}

for (county_name in names(county_contact_matrices)) {
  county_matrix <- county_contact_matrices[[county_name]]
  grade_group <- grade_groups_by_county[[county_name]]
  reference_eigenvalue <- 
    reference_eigenvalues[[which(sapply(grade_groups, identical, grade_group))]]
  R0_factor <- Re(eigen(county_matrix)$values[1]) / Re(reference_eigenvalue)
  county_R0_factors <- rbind(county_R0_factors, data.frame(
    county = county_name, R0_factor = R0_factor))
}

county_R0_factors$R0_rank <- rank(-county_R0_factors$R0_factor)
```

## Rv by county

```{r county-Rv}
get_Rv <- function(R0, popsize, contactMatrix, initImmune){
  vaxrepnum(1, 
            popsize,
            transmissionRates(R0, 1, contactMatrix),
            rep(0, length(popsize)),
            initImmune,
            1)
}

county_Rv <- rep(0, length(county_contact_matrices))
county_name <- names(county_contact_matrices)
for (i in seq_along(county_Rv)){
  county_matrix <- county_contact_matrices[[county_name[i]]]
  
  df <- county_df[county_df$county == county_name[i],]
  R0_local <- R0_baseline * 
    county_R0_factors$R0_factor[county_R0_factors$county == county_name[i]] 
  Rv <- get_Rv(R0_local,
               df$population,
               county_matrix,
               (df$MMR2 * 0.97 + 
                  df$MMR1 * 0.93 + 
                  df$priorinf) * df$population)
  county_Rv[i] <- Rv
}
county_Rvs <- data.frame(county = county_name, Rv = county_Rv)
county_Rvs$Rv_rank <- rank(-county_Rvs$Rv)
```

Create a table of county ranks

```{r county-rank-summary}
value_rank_summary <- merge(county_total_pop, merge(school_vax, merge(pop_immunity, county_Rvs)))

rank_summary <- value_rank_summary[,c(-3, -5, -7)]
```

