---
title: "Measles Age-Structured Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measles Age-Structured Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7
)
```

```{r setup}
library(multigroup.vaccine)
library(socialmixr)
library(ggplot2)

verbose <- FALSE  # Set to TRUE to enable detailed logging during processing
```

# Age-Structured UT County Model

This vignette is an experimental vignette in which we build and run an age-structured model of transmission within Utah on a county-level resolution.

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in this location.

## Age Structures

First, define the age groups for our model.

Because each county has a different school vax distribution, we need to use the same age groups as the school vax distribution for a given county.
In the vaccination data (county_vax_data.csv), the vaccination data is organized by **school** level (1=el, 2=middle, 3=high).

**Important methodological note on grade structure assumptions:**

School grade structures in Utah are determined at the **district level**, not county level. Within a single county, different districts may use different configurations (e.g., some K-5/6-8/9-12, others K-6/7-9/10-12). However, our vaccination data is aggregated at the **county level**.

For modeling purposes, we make the following **simplifying assumptions**:

1. **County-level aggregation is acceptable**: Since our vaccination data is county-aggregated, we treat each county as having a *dominant grade structure*
2. **Population ratios reveal dominant patterns**: We use the relative sizes of elementary/middle/high school populations to infer the most common grade configuration in that county
3. **Age groups approximate school levels**: Our age groups are designed to roughly align with school levels, acknowledging that exact grade-to-age mapping varies by district policies

This approach works best for counties with relatively uniform district structures.
Counties with highly heterogeneous district structures may require manual overrides based on demographic research.

### Heuristics

For the first (somewhat naive approach), we will use the *ratio* of the school populations to determine the age groups.

**Inference heuristic for age groups based on population ratios:**

The logic uses population ratios to infer grade structures:
- If middle ≈ 50% of high: Middle likely 2 grades (7-8) → elementary = K-6
- If middle ≈ 75% of high: Middle likely 3 grades (6-8) → elementary = K-5
- If middle ≈ high (ratio 80%-120%):
  - if elementary is much smaller than middle (ratio <0.5): likely K-4 for elementary, which supports 4-grade structures for middle and high (5-8, 9-12)
  - if elementary is larger than middle (ratio >0.5): likely K-6 for elementary, which supports 3-grade structures for middle and high (7-9, 10-12) (ie: Weber County fits this pattern)

**Limitations of this heuristic:**
- Assumes uniform grade structure across all districts within a county
- Does not account for charter schools or alternative school configurations
- Students enrolled in *online* schools may not be captured accurately in the population ratios
- Population ratios may be influenced by factors other than grade spans (e.g., demographic patterns, school choice)
- Manual overrides can be added for counties where specific district information is known (for example, Millard County has a known K-4/5-8/9-12 structure based on district data)

**Manual overrides:** For counties with verified grade structures that differ from the heuristic inference, add them to the `manual_overrides` list in the code below.

```{r define-age-groups}
county_vax_data <- read.csv(file.path("../../data-raw/county_vax_data.csv"))

age_groups_by_county <- list()

# Get unique counties
unique_counties <- unique(county_vax_data$county)

for (county_name in unique_counties) {
  # Extract data for this county
  county_data <- county_vax_data[county_vax_data$county == county_name, ]
  
  # Get population for each level
  el_pop <- county_data$pop[county_data$level == 1]
  middle_pop <- county_data$pop[county_data$level == 2]
  high_pop <- county_data$pop[county_data$level == 3]
  
  if (verbose) {
    print(paste("Processing county:", county_name))
    print(paste("Elementary pop:", el_pop, "Middle pop:", middle_pop, "High pop:", high_pop))
  }
  
  # Some counties have more unique structures, so we can define manual overrides for those based on known district configurations
  # This is especially important for rural counties.
  manual_overrides <- list(
    # Example:
    # "Millard" = list(
    #   ages = c(0, 5, 10, 14, 18, 25, 45, 70),
    #   reason = "K-4 elementary, 5-8 middle, 9-12 high (verified from district data)"
    # )
  )

  # Check for manual override first
  if (county_name %in% names(manual_overrides)) {
    override <- manual_overrides[[county_name]]
    age_groups_by_county[[county_name]] <- override$ages
    if (verbose) {
      print(paste("  -> Using manual override:", override$reason))
      print(paste("     Ages:", paste(override$ages, collapse=", ")))
    }
  } else if (length(middle_pop) > 0 && length(high_pop) > 0 && high_pop > 0) {
    # Use population ratio to determine age structure
    middle_to_high_ratio <- middle_pop / high_pop
    el_to_middle_ratio <- if(length(el_pop) > 0 && middle_pop > 0) el_pop / middle_pop else NA

    if (verbose) {
      print(paste("Middle to high ratio:", round(middle_to_high_ratio, 3)))
      if (!is.na(el_to_middle_ratio)) {
        print(paste("Elementary to middle ratio:", round(el_to_middle_ratio, 3)))
      }
    }

    # logic to detect different grade structures
    if (middle_to_high_ratio >= 0.8 && middle_to_high_ratio <= 1.2) {  # 1.2 is used to provide 'padding' for small population noise
      # if the elementary to middle ratio is low (e.g., <1.3), it may indicate a K-4 structure for elementary, which would further support the 4-grade inference for middle and high
      # however, if the elementary to middle ratio is higher (e.g., >1.3), it may indicate a K-6 structure for elementary, which would suggest a 3-grade structure for middle (7-9) and 3-grade for high (10-12)
      if (!is.na(el_to_middle_ratio) && el_to_middle_ratio < 1.3) {
        # Elementary is much smaller than middle → likely K-4
        # Suggests: Elementary K-4, Middle 5-8, High 9-12
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 10, 14, 18, 25, 45, 70)
        if (verbose) print("  -> Using ages: 0, 1, 5, 10, 14, 18, 25, 45, 70 (inferred: el=K-4, middle=5-8, high=9-12)")
      } else {
        # Elementary is larger than middle → likely K-6
        # Suggests: Elementary K-5, Middle 7-9, High 10-12
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 15, 18, 25, 45, 70)
        if (verbose) print("  -> Using ages: 0, 1, 5, 12, 15, 18, 25, 45, 70 (inferred: el=K-5, middle=7-9, high=10-12)")
      }
    } else if (middle_to_high_ratio <= 0.6) {
      # Middle is ~50% of high → middle likely 2 grades (7-8)
      # Suggests: Elementary K-6, Middle 7-8, High 9-12
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
      if (verbose) print("  -> Using ages: 0, 1, 5, 12, 14, 18, 25, 45, 70 (inferred: el=K-6, middle=7-8, high=9-12)")
    } else {
      # Middle is ~60-80% of high → middle likely 3 grades (6-8)
      # Suggests: Elementary K-5, Middle 6-8, High 9-12
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
      if (verbose) print("  -> Using ages: 0, 1, 5, 11, 14, 18, 25, 45, 70 (inferred: el=K-5, middle=6-8, high=9-12)")
    }
  } else {
    # Default if data is missing
    age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
    if (verbose) print("  -> Using default ages: 0, 1, 5, 11, 14, 18, 25, 45, 70")
  }
  if (verbose) print("")
}

if (verbose) {
  print("Age groups by county:")
  print(age_groups_by_county)
  
  print("Number of counties processed:")
  print(length(age_groups_by_county))
}

# One of the 'counties' is Online, which represents online schools in UT.
# Remove that county from the list and save it as a separate variable
online_schools_age_groups <- age_groups_by_county$Online
age_groups_by_county$Online <- NULL

if (verbose) {
  print("Age groups for online schools:")
  print(online_schools_age_groups)
}
```

Now get all counties from Utah, passing in the age groups for each county as defined above:

```{r get-all-counties}
# get all counties from Utah using getAllCountiesData with county-specific age groups
# data is pre-saved as csv in data-raw
utah_counties <- getAllCountiesData(
  state_fips = getStateFIPS("Utah"),
  year = 2024,
  age_groups_by_county = age_groups_by_county,
  csv_path = getCensusDataPath(),
  verbose = verbose
)

# Display the names of all counties
names(utah_counties)

# Show example data from one county
if (verbose) {
  print("Example county data:")
  print(utah_counties[[1]])
}
```

```{r visualize-age-structures}
# Categorize each county by its age group structure
# Extract the school-age portions (indices 3-5) to identify the structure type
structure_df <- data.frame(
  county = character(),
  structure_type = character(),
  el_start = numeric(),
  el_end = numeric(),
  middle_start = numeric(),
  middle_end = numeric(),
  high_start = numeric(),
  high_end = numeric(),
  stringsAsFactors = FALSE
)

for (county_name in names(age_groups_by_county)) {
  ages <- age_groups_by_county[[county_name]]

  # School age groups are typically at indices 3, 4, 5, 6
  # (after 0, 1, and the first school boundary)
  # For visualization, we want segments to connect, so end points should equal the next start point
  el_start <- ages[3]
  el_end <- ages[4]  # This connects to middle_start
  middle_start <- ages[4]
  middle_end <- ages[5]  # This connects to high_start
  high_start <- ages[5]
  high_end <- ages[6]  # This is the final boundary

  # Categorize based on the age structure
  if (identical(ages, c(0, 1, 5, 10, 14, 18, 25, 45, 70))) {
    structure_type <- "K-4 / 5-8 / 9-12"
  } else if (identical(ages, c(0, 1, 5, 12, 15, 18, 25, 45, 70))) {
    structure_type <- "K-6 / 7-9 / 10-12"
  } else if (identical(ages, c(0, 1, 5, 12, 14, 18, 25, 45, 70))) {
    structure_type <- "K-6 / 7-8 / 9-12"
  } else if (identical(ages, c(0, 1, 5, 11, 14, 18, 25, 45, 70))) {
    structure_type <- "K-5 / 6-8 / 9-12"
  } else {
    structure_type <- "Other"
  }

  structure_df <- rbind(structure_df, data.frame(
    county = gsub(" County", "", county_name),
    structure_type = structure_type,
    el_start = el_start,
    el_end = el_end,
    middle_start = middle_start,
    middle_end = middle_end,
    high_start = high_start,
    high_end = high_end
  ))
}

# Count by structure type
structure_counts <- table(structure_df$structure_type)
print("Distribution of grade structures:")
print(structure_counts)

# Create a visual showing the grade spans for each county
library(tidyr)
library(dplyr)

# Reshape data for plotting - create one row per county per school level
structure_long <- rbind(
  structure_df %>% select(county, structure_type, start = el_start, end = el_end) %>% mutate(level = "Elementary"),
  structure_df %>% select(county, structure_type, start = middle_start, end = middle_end) %>% mutate(level = "Middle"),
  structure_df %>% select(county, structure_type, start = high_start, end = high_end) %>% mutate(level = "High")
)

# Order counties by structure type and then alphabetically
structure_long <- structure_long %>%
  arrange(structure_type, county) %>%
  mutate(county = factor(county, levels = unique(county)))

# Plot showing grade spans for each county
ggplot(structure_long, aes(y = county)) +
  geom_segment(aes(x = start, xend = end, yend = county, color = level), 
               size = 6, alpha = 0.8) +
  scale_color_manual(values = c("Elementary" = "#1f77b4", 
                                  "Middle" = "#ff7f0e", 
                                  "High" = "#2ca02c")) +
  scale_x_continuous(breaks = seq(0, 20, by = 1), expand = c(0, 0)) +
  facet_grid(structure_type ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Grade Structure Age Spans by County",
    subtitle = "Counties grouped by inferred grade configuration",
    x = "Age",
    y = "County",
    color = "School Level"
  ) +
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0, hjust = 0),
    legend.position = "bottom"
  )
```

### Difference between Vax and Census data

There is a non-negligible difference between the census 2024 estimate age-grouping population sizes and the population sizes from the vaccination data.

This could be due to a variety of external factors.

To understand this difference, we can compare the processed age-groupings and the population sizes from the vaccination data for every county.

```{r compare-vax-census}
county_difference <- data.frame(
  county = character(),
  census_population = numeric(),
  vax_population = numeric(),
  difference = numeric(),
  stringsAsFactors = FALSE
)

# For every county, compute the difference at each age group between the census population and the vaccination data population
for (county_name in names(utah_counties)) {
  census_data <- utah_counties[[county_name]]

  # Remove " County" from the county name to match the format in the vaccination data
  county_name <- gsub(" County", "", county_name)
  vax_data <- county_vax_data[county_vax_data$county == county_name, ]

  if (verbose) print(paste("County:", county_name))

  # Get the population sizes for the school age groups from the census data and the vaccination data
  census_age_1 <- census_data$age_labels[3]  # age group corresponding to the first school age group (e.g., 1-4 or 1-5)
  census_age_2 <- census_data$age_labels[4]  # age group corresponding to the second school age group (e.g., 5-11 or 6-11)
  census_age_3 <- census_data$age_labels[5]  # age group corresponding to the third school age group (e.g., 12-17)

  census_pop <- census_data$age_pops[census_data$age_labels %in% c(census_age_1, census_age_2, census_age_3)]

  vax_pop <- c(vax_data$pop[vax_data$level == 1], vax_data$pop[vax_data$level == 2], vax_data$pop[vax_data$level == 3])

  difference <- sum(census_pop) - sum(vax_pop)

  county_difference <- rbind(county_difference, data.frame(
    county = county_name,
    census_population = sum(census_pop),
    vax_population = sum(vax_pop),
    difference = difference
  ))

  if (verbose) {
    print(paste("County:", county_name))
    print(paste("Census population:", sum(census_pop), "Vax population:",
                sum(vax_pop), "Difference:", sum(census_pop) - sum(vax_pop)))
    print("")
  }
}
```

```{r visualize-age-differences}
library(scales)  # For comma formatter
print(county_difference)

county_difference$total_population <- county_difference$census_population
county_difference$percent_difference <- abs((county_difference$difference / county_difference$census_population) * 100)

# Bar chart: absolute difference in population size between census and vaccination data for each county
ggplot(county_difference, aes(x = reorder(county, difference), y = difference)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(round(percent_difference, 1), "%")), vjust = -0.5, size = 3) +
  labs(
    title = "Difference in Population Size between Census and Vaccination Data by County",
    subtitle = "Labels show percentage difference",
    x = "County",
    y = "Difference in Population Size"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Two-panel view showing both absolute and percentage differences
library(gridExtra)
p1 <- ggplot(county_difference, aes(x = total_population, y = difference)) +
  geom_point(color = "darkblue", size = 3, alpha = 0.7) +
  geom_text(aes(label = county), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Difference",
    x = "Census School-Age Population (log scale)",
    y = "Difference (Census - Vax)"
  ) +
  theme_minimal()

p2 <- ggplot(county_difference, aes(x = total_population, y = percent_difference)) +
  geom_point(color = "darkred", size = 3, alpha = 0.7) +
  geom_text(aes(label = county), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Percentage Difference",
    x = "Census School-Age Population (log scale)",
    y = "Percentage Difference"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 1)
```

### Vaccination Levels by Age Group

Now we assign vaccination/immunity levels to each age group for each county based on the school vaccination data. We follow the approach from the measles age model vignette, accounting for:

- Under 1 year: No vaccination (0%)
- Ages 1-4: Slightly lower than elementary (90% of elementary rate, with 93% vaccine effectiveness for single dose)
- School-age groups: Use the actual vaccination rates from the data (with 97% vaccine effectiveness)
- Adults 18-69: linearly extrapolate from high school rates
- Ages 70+: Very high immunity (99%)

```{r vaccination-levels}
# Create a data frame to store immunity levels for each county
county_immunity <- list()

for (county_name in names(utah_counties)) {
  # Get the age groups for this county
  county_data <- utah_counties[[county_name]]
  county_name_short <- gsub(" County", "", county_name)
  
  # Get vaccination data for this county
  vax_data <- county_vax_data[county_vax_data$county == county_name_short, ]
  
  # Calculate vaccination rates directly from the data for each school level
  # These are the actual vaccination rates: vax / pop
  el_vax_rate <- vax_data$vax[vax_data$level == 1] / vax_data$pop[vax_data$level == 1]
  middle_vax_rate <- vax_data$vax[vax_data$level == 2] / vax_data$pop[vax_data$level == 2]
  high_vax_rate <- vax_data$vax[vax_data$level == 3] / vax_data$pop[vax_data$level == 3]
  pre_k_vax_rate <- el_vax_rate * 0.9  # Assuming pre-K has about 90% of the elementary vaccination rate

  # Calculate immunity by applying vaccine effectiveness
  # School age: 97% effectiveness (two-dose schedule)
  # Ages 1-4: 93% effectiveness (single-dose)
  pre_k_immunity <- pre_k_vax_rate * 0.93
  el_immunity <- el_vax_rate * 0.97
  middle_immunity <- middle_vax_rate * 0.97
  high_immunity <- high_vax_rate * 0.97

  # Get the age groups for this county to determine midpoints for interpolation
  ages <- age_groups_by_county[[county_name]]

  # Linear interpolation from high school immunity at age 18 to 100% at age 70
  # Starting point: high school immunity at age 18
  start_immunity <- high_immunity
  end_immunity <- 1.0  # Target 100% at age 70
  start_age <- 18
  end_age <- 70

  # Calculate midpoint ages for adult groups
  age_18_25_mid <- (18 + 25) / 2  # 21.5
  age_25_45_mid <- (25 + 45) / 2  # 35
  age_45_70_mid <- (45 + 70) / 2  # 57.5

  # Linear interpolation function
  interpolate_immunity <- function(age, start_age, end_age, start_imm, end_imm) {
    slope <- (end_imm - start_imm) / (end_age - start_age)
    return(start_imm + slope * (age - start_age))
  }

  # Calculate immunity for adult groups
  immunity_18_25 <- interpolate_immunity(age_18_25_mid, start_age, end_age, start_immunity, end_immunity)
  immunity_25_45 <- interpolate_immunity(age_25_45_mid, start_age, end_age, start_immunity, end_immunity)
  immunity_45_70 <- interpolate_immunity(age_45_70_mid, start_age, end_age, start_immunity, end_immunity)

  # Build the age_immunity vector based on the county's age groups
  # The structure is: [0-1, 1-4/5, elementary, middle, high, 18-25, 25-45, 45-70, 70+]
  age_immunity <- c(
    0,                                    # Under 1: no vaccination
    pre_k_immunity,                       # 1-4: 90% of elementary vax rate, with 93% effectiveness
    el_immunity,                          # Elementary: from vax data (level 1)
    middle_immunity,                      # Middle: from vax data (level 2)
    high_immunity,                        # High: from vax data (level 3)
    immunity_18_25,                       # 18-24: linearly interpolated from high school
    immunity_25_45,                       # 25-44: linearly interpolated
    immunity_45_70,                       # 45-69: linearly interpolated
    end_immunity                          # 70+: 100% immunity
  )

  county_immunity[[county_name]] <- age_immunity

  if (verbose) {
    print(paste("County:", county_name_short))
    print(paste("  Elementary vax rate:", round(el_vax_rate, 3), "-> immunity:", round(el_immunity, 3)))
    print(paste("  Middle vax rate:", round(middle_vax_rate, 3), "-> immunity:", round(middle_immunity, 3)))
    print(paste("  High vax rate:", round(high_vax_rate, 3), "-> immunity:", round(high_immunity, 3)))
    print("  Age immunity vector:")
    print(round(age_immunity, 3))
    print("")
  }
}
```

## R0 Calibration Based on Age Distribution

Before estimating outbreak sizes, we calibrate R0 values for each county based on their specific age distributions and contact patterns. Following the approach from the age distribution R0 analysis, we calculate adjustment factors based on contact matrix eigenvalues.

**Methodology:**

- Create a reference contact matrix using default Polymod survey data
- Calculate the dominant eigenvalue of each county's contact matrix
- Compute R0 adjustment factors relative to the reference
- Apply these factors to obtain county-specific effective R0 values

**Assumptions:**

- Baseline R0 for measles is set at 10
- Contact patterns are assumed to be similar to those captured in the Polymod survey

```{r R0-calibration}
# Baseline R0 for measles
R0_baseline <- 10

# Create reference contact matrix using default Polymod data
# Using a common age structure for reference
reference_age_limits <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
reference_matrix <- contactMatrixPolymod(reference_age_limits)
reference_eigenvalue <- eigen(reference_matrix)$values[1]

if (verbose) {
  print(paste("Reference eigenvalue:", round(Re(reference_eigenvalue), 3)))
}

# Calculate R0 adjustment factors for each county
county_R0_factors <- data.frame(
  county = character(),
  eigenvalue = numeric(),
  R0_factor = numeric(),
  R0_effective = numeric(),
  stringsAsFactors = FALSE
)

for (county_name in names(utah_counties)) {
  county_name_short <- gsub(" County", "", county_name)
  county_data <- utah_counties[[county_name]]
  age_limits <- age_groups_by_county[[county_name_short]]
  popsize <- county_data$age_pops

  # Create county-specific contact matrix
  county_matrix <- contactMatrixPolymod(age_limits, popsize)

  # Calculate dominant eigenvalue
  county_eigenvalue <- eigen(county_matrix)$values[1]

  # Calculate R0 adjustment factor
  R0_factor <- Re(county_eigenvalue) / Re(reference_eigenvalue)

  # Calculate effective R0 for this county
  R0_effective <- R0_baseline * R0_factor

  county_R0_factors <- rbind(county_R0_factors, data.frame(
    county = county_name_short,
    eigenvalue = Re(county_eigenvalue),
    R0_factor = R0_factor,
    R0_effective = R0_effective
  ))

  if (verbose) {
    print(paste("County:", county_name_short))
    print(paste("  Eigenvalue:", round(Re(county_eigenvalue), 3)))
    print(paste("  R0 factor:", round(R0_factor, 3)))
    print(paste("  Effective R0:", round(R0_effective, 2)))
  }
}

# Summary of R0 calibration
print("R0 Calibration Summary:")
print(paste("Baseline R0:", R0_baseline))
print(paste("Mean R0 factor:", round(mean(county_R0_factors$R0_factor), 3)))
print(paste("Range of R0 factors:", round(min(county_R0_factors$R0_factor), 3), 
            "to", round(max(county_R0_factors$R0_factor), 3)))
print(paste("Range of effective R0:", round(min(county_R0_factors$R0_effective), 2),
            "to", round(max(county_R0_factors$R0_effective), 2)))
```

```{r visualize-R0-calibration}
# Histogram of R0 factors
ggplot(county_R0_factors, aes(x = R0_factor)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "Distribution of R0 Adjustment Factors by County",
    subtitle = "Factors > 1 indicate age distributions more favorable to transmission",
    x = "R0 Adjustment Factor",
    y = "Number of Counties"
  ) +
  theme_minimal()

# Effective R0 by county
ggplot(county_R0_factors, aes(x = reorder(county, R0_effective), y = R0_effective)) +
  geom_point(size = 3, alpha = 0.7, color = "darkblue") +
  geom_hline(yintercept = R0_baseline, linetype = "dashed", color = "red", 
             alpha = 0.5) +
  coord_flip() +
  labs(
    title = "Effective R0 by County (Age-Adjusted)",
    subtitle = paste0("Based on baseline R0 = ", R0_baseline, " (red line)"),
    x = "County",
    y = "Effective R0"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7))
```

### Sensitivity to Baseline R0

Changing the baseline R0 assumption will affect the effective R0 values for each county.

To understand this sensitivity, we can test a range of baseline R0 values from 6 to 16.
```{r R0-sensitivity-analysis}
# Test range of baseline R0 values
R0_baseline_range <- seq(6, 16, by = 1)

# Create data frame to store results across all baselines and counties
R0_sensitivity_data <- data.frame()

for (R0_test in R0_baseline_range) {
  for (i in 1:nrow(county_R0_factors)) {
    county_name <- county_R0_factors$county[i]
    R0_factor <- county_R0_factors$R0_factor[i]
    R0_eff <- R0_test * R0_factor

    # Calculate herd immunity threshold for this effective R0
    # TODO: update to use vaxrepnum rather than naive calculation
    HIT <- 1 - (1 / R0_eff)

    R0_sensitivity_data <- rbind(R0_sensitivity_data, data.frame(
      baseline_R0 = R0_test,
      county = county_name,
      R0_factor = R0_factor,
      effective_R0 = R0_eff,
      herd_immunity_threshold = HIT
    ))
  }
}

# Summary statistics by baseline R0
sensitivity_summary <- R0_sensitivity_data %>%
  group_by(baseline_R0) %>%
  summarise(
    min_effective_R0 = min(effective_R0),
    mean_effective_R0 = mean(effective_R0),
    max_effective_R0 = max(effective_R0),
    mean_HIT = mean(herd_immunity_threshold)
  )

print("Sensitivity Analysis Summary:")
print(sensitivity_summary)
```

```{r visualize-R0-sensitivity}
# Heatmap: Effective R0 by county and baseline
ggplot(R0_sensitivity_data, aes(x = baseline_R0, y = reorder(county, effective_R0), fill = effective_R0)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "yellow", high = "red", 
                       midpoint = 10, name = "Effective R0") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Effective R0 Heatmap: County vs Baseline R0",
    subtitle = "Shows how age-adjusted R0 varies with baseline assumptions",
    x = "Baseline R0",
    y = "County"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

# Line plot: Effective R0 trajectories for each county
ggplot(R0_sensitivity_data, aes(x = baseline_R0, y = effective_R0, group = county)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0.02, 0)) +
  labs(
    title = "Effective R0 Across Baseline Values",
    subtitle = "Individual counties (blue) vs 1:1 ratio (red dashed line)",
    x = "Baseline R0",
    y = "Effective R0"
  ) +
  theme_minimal()

# Heatmap: Herd immunity threshold by county and baseline
ggplot(R0_sensitivity_data, aes(x = baseline_R0, y = reorder(county, herd_immunity_threshold), 
                                 fill = herd_immunity_threshold * 100)) +
  geom_tile() +
  scale_fill_gradient2(low = "green", mid = "yellow", high = "red", 
                       midpoint = 90, name = "HIT (%)") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Herd Immunity Threshold Heatmap: County vs Baseline R0",
    subtitle = "Critical vaccination coverage needed varies by age structure",
    x = "Baseline R0",
    y = "County"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

# Violin plot: Distribution of effective R0 at each baseline (raincloud style)
# One-sided KDE with jittered points inside and boxplot, all aligned on same axis
library(ggdist)

R0_sensitivity_data$baseline_R0_factor <- factor(R0_sensitivity_data$baseline_R0)

ggplot(R0_sensitivity_data, aes(y = baseline_R0_factor, x = effective_R0)) +
  # right side KDE
  stat_halfeye(aes(group = baseline_R0_factor),
               side = "right",
               fill = "steelblue", alpha = 0.5,
               .width = 0, point_interval = NULL,
               scale = 0.8) +
  # Boxplot - left side
  geom_boxplot(aes(group = baseline_R0_factor),
               width = 0.2, outlier.shape = NA, alpha = 0.7, fill = "gray80") +
  # Jittered points - centered on the axis with slight vertical jitter for visibility
  geom_jitter(aes(group = baseline_R0_factor),
              width = 0, height = 0.1, alpha = 0.4, size = 1.5, color = "darkblue") +
  labs(
    title = "Distribution of Effective R0 Across Counties",
    subtitle = "Boxplot (left), individual counties (center), and KDE (right) aligned on discrete baseline R0",
    y = "Baseline R0",
    x = "Effective R0"
  ) +
  theme_minimal()
```

## Outbreak size estimation for every county

Now we use the `finalsize()` function to estimate outbreak size for each county, using:
- Age-specific population sizes and immunity levels
- County-specific calibrated R0 values adjusted for age distribution (assumes baseline R0 of 10)

**Note:** Hybrid model would provide a better method for risk determination
TODO: convert to hybrid model

```{r outbreak-estimation}
# Store results for all counties
outbreak_results <- data.frame(
  county = character(),
  total_outbreak_size = numeric(),
  population = numeric(),
  susceptible_population = numeric(),
  R0_effective = numeric(),
  herd_immunity_threshold = numeric(),
  stringsAsFactors = FALSE
)

# Store age-specific results
outbreak_by_age <- list()

# Transmission parameters
relsusc <- NULL  # Will be set based on number of age groups per county
reltransm <- NULL  # Will be set based on number of age groups per county

for (county_name in names(utah_counties)) {
  county_data <- utah_counties[[county_name]]
  county_name_short <- gsub(" County", "", county_name)

  if (verbose) {
    print(paste("Processing outbreak estimation for:", county_name_short))
  }

  # Get population and immunity data
  popsize <- county_data$age_pops
  age_immunity <- county_immunity[[county_name]]
  age_limits <- age_groups_by_county[[county_name_short]]

  # Get county-specific calibrated R0
  R0_effective <- county_R0_factors$R0_effective[county_R0_factors$county == county_name_short]

  if (verbose) {
    print(paste("  Age limits length:", length(age_limits)))
    print(paste("  Population size length:", length(popsize)))
    print(paste("  Age immunity length:", length(age_immunity)))
    print(paste("  Effective R0:", round(R0_effective, 2)))
  }

  # Initial conditions
  initV <- round(age_immunity * popsize)  # initially immune individuals
  initI <- rep(0, length(popsize))        # initial infectious cases
  initI[6] <- 1                           # assume 1 initial case in 6th age group (young adults)
  initR <- rep(0, length(popsize))        # no recent prior outbreaks

  # Set up transmission parameters
  relsusc <- rep(1, length(popsize))      # no age differences in susceptibility
  reltransm <- rep(1, length(popsize))    # or transmissibility per contact

  # Create contact matrix using Polymod data
  contactmatrix <- contactMatrixPolymod(age_limits, popsize)

  # Run finalsize model with calibrated R0
  fs <- finalsize(popsize, R0_effective, contactmatrix, relsusc, reltransm, initR, initI, initV)

  # Store results
  total_cases <- sum(fs)
  total_pop <- sum(popsize)
  susceptible_pop <- total_pop - sum(initV)

  # Calculate herd immunity threshold: proportion that needs to be immune = 1 - 1/R0
  # TODO: update to use vaxrepnum rather than naive calculation
  herd_immunity_threshold <- 1 - (1 / R0_effective)

  outbreak_results <- rbind(outbreak_results, data.frame(
    county = county_name_short,
    total_outbreak_size = total_cases,
    population = total_pop,
    susceptible_population = susceptible_pop,
    R0_effective = R0_effective,
    herd_immunity_threshold = herd_immunity_threshold
  ))

  # Store age-specific results
  age_specific_df <- data.frame(
    age_group = county_data$age_labels,
    cases = fs,
    population = popsize,
    susceptible = popsize - initV[1:length(popsize)]
  )
  outbreak_by_age[[county_name_short]] <- age_specific_df

  if (verbose) {
    print(paste("  Total outbreak size:", round(total_cases)))
    print(paste("  Susceptible population:", round(susceptible_pop)))
    print(paste("  Herd immunity threshold:", round(herd_immunity_threshold * 100, 1), "%"))
  }
}

# Sort by outbreak size
outbreak_results <- outbreak_results[order(-outbreak_results$total_outbreak_size), ]

if (verbose) {
  print("Outbreak estimation complete for all counties")
  print(head(outbreak_results))
}
```

```{r visualize-outbreak-results}
# Calculate actual immunity coverage and compare to threshold
outbreak_results$actual_immunity <- 1 - (outbreak_results$susceptible_population / outbreak_results$population)
outbreak_results$immunity_gap <- outbreak_results$herd_immunity_threshold - outbreak_results$actual_immunity
outbreak_results$final_size_percent <- (outbreak_results$total_outbreak_size / outbreak_results$population) * 100

# Bar chart: Total outbreak size by county (absolute cases)
ggplot(outbreak_results, aes(x = reorder(county, total_outbreak_size), y = total_outbreak_size)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Estimated Final Epidemic Size by Utah County",
    subtitle = "Using age-distribution calibrated R0 and one initial case in young adults",
    x = "County",
    y = "Final Epidemic Size (total cases)"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Bar chart: Final epidemic size as percentage of population
ggplot(outbreak_results, aes(x = reorder(county, final_size_percent), y = final_size_percent)) +
  geom_bar(stat = "identity", fill = "darkred", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Final Epidemic Size by Utah County (Percentage)",
    subtitle = "Cumulative proportion infected by end of epidemic",
    x = "County",
    y = "Final Epidemic Size (% of population)"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Create summary statistics
summary_stats <- data.frame(
  Statistic = c(
    "Baseline R0",
    "Mean Effective R0",
    "Mean Herd Immunity Threshold (%)",
    "Mean Actual Immunity (%)", 
    "Median Actual Immunity (%)",
    "Counties Below Threshold",
    "Mean Final Epidemic Size (%)",
    "Median Final Epidemic Size (%)",
    "Total Cases (all counties)", 
    "Total Population (all counties)",
    "Total Susceptible (all counties)"
  ),
  Value = c(
    R0_baseline,
    round(mean(outbreak_results$R0_effective), 2),
    round(mean(outbreak_results$herd_immunity_threshold) * 100, 2),
    round(mean(outbreak_results$actual_immunity) * 100, 2),
    round(median(outbreak_results$actual_immunity) * 100, 2),
    sum(outbreak_results$immunity_gap > 0),
    round(mean(outbreak_results$final_size_percent), 2),
    round(median(outbreak_results$final_size_percent), 2),
    round(sum(outbreak_results$total_outbreak_size)),
    sum(outbreak_results$population),
    round(sum(outbreak_results$susceptible_population))
  )
)

print("Summary Statistics:")
print(summary_stats)
```
