---
title: "Measles Risk Assessment: Vulnerability & Consequence in Utah Counties"
date: "March 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measles Risk Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

Executive Summary

This assessment evaluates the risk of a measles outbreak across Utah counties. Traditional risk assessments rely strictly on raw school vaccination rates. However, this approach ignores a critical factor: age-structured contact patterns.

Using a stochastic epidemiological model, this analysis combines demographic mixing patterns with local vaccination coverage to estimate two specific metrics for every county:

    Vulnerability (Rv​): The vaccination reproduction number. If Rv​≥1, the county is vulnerable to sustained transmission. If Rv​<1, the county has achieved herd immunity.

    Consequence: If an outbreak is introduced, what is the probability it grows into a major event (>100 cases), and what is the expected scale of the outbreak in absolute numbers?

1. Data Initialization & Demographic Calibration

Note: This section processes county-level census data, school vaccination records, and age-structured contact matrices (Polymod). The baseline pathogen transmissibility (R0​) is calibrated to account for the unique age distribution of each county.

```{r setup-libs, cache=FALSE}
library(multigroup.vaccine)
library(socialmixr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)
library(scales)

# Load raw data
county_vax_data <- read.csv(file.path("../../data-raw/county_vax_data.csv"))
unique_counties <- unique(county_vax_data$county)

# Infer age structures based on population ratios
age_groups_by_county <- list()
for (county_name in unique_counties) {
  county_data <- county_vax_data[county_vax_data$county == county_name, ]
  el_pop <- county_data$pop[county_data$level == 1]
  middle_pop <- county_data$pop[county_data$level == 2]
  high_pop <- county_data$pop[county_data$level == 3]
  
  if (length(middle_pop) > 0 && length(high_pop) > 0 && high_pop > 0) {
    middle_to_high_ratio <- middle_pop / high_pop
    el_to_middle_ratio <- if(length(el_pop) > 0 && middle_pop > 0) el_pop / middle_pop else NA

    if (middle_to_high_ratio >= 0.8 && middle_to_high_ratio <= 1.2) {
      if (!is.na(el_to_middle_ratio) && el_to_middle_ratio < 1.3) {
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 10, 14, 18, 25, 45, 70)
      } else {
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 15, 18, 25, 45, 70)
      }
    } else if (middle_to_high_ratio <= 0.6) {
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
    } else {
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
    }
  } else {
    age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
  }
}
age_groups_by_county$Online <- NULL

# Fetch Census Data
utah_counties <- getAllCountiesData(
  state_fips = getStateFIPS("Utah"),
  year = 2024,
  age_groups_by_county = age_groups_by_county,
  csv_path = getCensusDataPath(),
  verbose = FALSE
)

# Calculate Immunity Levels & Calibrate R0
R0_baseline <- 10
R0_baseline_range <- seq(6, 16, by=1)
reference_matrix <- contactMatrixPolymod(c(0, 1, 5, 11, 14, 18, 25, 45, 70))
reference_eigenvalue <- eigen(reference_matrix)$values[1]

county_immunity <- list()
county_contact_matrices <- list()
county_R0_factors <- data.frame(county = character(), R0_factor = numeric(), R0_effective = numeric())

for (county_name in names(utah_counties)) {
  county_name_short <- gsub(" County", "", county_name)
  county_data <- utah_counties[[county_name]]
  vax_data <- county_vax_data[county_vax_data$county == county_name_short, ]
  
  el_vax_rate <- vax_data$vax[vax_data$level == 1] / vax_data$pop[vax_data$level == 1]
  middle_vax_rate <- vax_data$vax[vax_data$level == 2] / vax_data$pop[vax_data$level == 2]
  high_vax_rate <- vax_data$vax[vax_data$level == 3] / vax_data$pop[vax_data$level == 3]
  pre_k_vax_rate <- el_vax_rate * 0.9  
  
  pre_k_immunity <- pre_k_vax_rate * 0.93
  el_immunity <- el_vax_rate * 0.97
  middle_immunity <- middle_vax_rate * 0.97
  high_immunity <- high_vax_rate * 0.97

  interpolate_immunity <- function(age, start_age, end_age, start_imm, end_imm) {
    slope <- (end_imm - start_imm) / (end_age - start_age)
    return(start_imm + slope * (age - start_age))
  }
  
  imm_18_25 <- interpolate_immunity(21.5, 18, 70, high_immunity, 1.0)
  imm_25_45 <- interpolate_immunity(35, 18, 70, high_immunity, 1.0)
  imm_45_70 <- interpolate_immunity(57.5, 18, 70, high_immunity, 1.0)

  county_immunity[[county_name]] <- c(0, pre_k_immunity, el_immunity, middle_immunity, high_immunity, imm_18_25, imm_25_45, imm_45_70, 1.0)
  
  # Contact Matrix & Eigenvalue adjustment
  county_matrix <- contactMatrixPolymod(age_groups_by_county[[county_name_short]], county_data$age_pops)
  county_contact_matrices[[county_name_short]] <- county_matrix
  
  R0_factor <- Re(eigen(county_matrix)$values[1]) / Re(reference_eigenvalue)
  county_R0_factors <- rbind(county_R0_factors, data.frame(
    county = county_name_short, R0_factor = R0_factor, R0_effective = R0_baseline * R0_factor
  ))
}
```

2. The Age-Structured Reality (Rv​)

A key question for public health allocation is whether incorporating age-structured contact patterns provides meaningfully different risk rankings compared to simply ranking counties by their raw school vaccination rates.

If the rankings diverge, it indicates that looking at vaccination rates alone causes us to underestimate the risk in certain counties due to their specific demographic mixing.

```{r rv-rankings, cache=TRUE}
# Compute overall school vaccination rate per county
school_vax_summary <- county_vax_data %>%
  filter(county != "Online") %>%
  group_by(county) %>%
  summarise(
    total_enrolled = sum(pop),
    school_vax_rate = sum(vax) / sum(pop),
    .groups = "drop"
  )

# Calculate Rv and rankings for each baseline R0
comparison_all <- data.frame()
for (R0_test in R0_baseline_range) {
  comparison_df <- merge(school_vax_summary, county_R0_factors, by = "county")
  comparison_df$vax_repnum <- sapply(comparison_df$county, function(c_name) {
    c_full <- paste0(c_name, " County")
    popsize <- utah_counties[[c_full]]$age_pops
    initV <- round(county_immunity[[c_full]] * popsize)
    R0_eff <- R0_test * county_R0_factors$R0_factor[county_R0_factors$county == c_name]
    trmat <- transmissionRates(R0_eff, 1, county_contact_matrices[[c_name]])
    Re(vaxrepnum(1, popsize, trmat, rep(0, length(popsize)), initV, 1))
  })
  comparison_df$rank_vax_rate <- rank(comparison_df$school_vax_rate)
  comparison_df$rank_Rv <- rank(-comparison_df$vax_repnum)
  comparison_df$rank_diff <- comparison_df$rank_vax_rate - comparison_df$rank_Rv
  comparison_df$baseline_R0 <- R0_test
  comparison_all <- rbind(comparison_all, comparison_df)
}

# Interactive rank difference chart with R0 slider
rank_order <- comparison_all %>% filter(baseline_R0 == 10) %>% arrange(rank_diff) %>% pull(county)

p_rank <- plot_ly()
for (r0_val in R0_baseline_range) {
  df_sub <- comparison_all %>% filter(baseline_R0 == r0_val)
  df_sub$county_f <- factor(df_sub$county, levels = rank_order)
  df_sub <- df_sub %>% arrange(county_f)

  p_rank <- p_rank %>% add_trace(
    data = df_sub, x = ~rank_diff, y = ~county_f, type = "bar", orientation = "h",
    marker = list(color = ifelse(df_sub$rank_diff > 0, "#d62728", "#2ca02c")),
    hovertext = ~paste0(county, "<br>Rank Diff: ", rank_diff, "<br>Rv: ", round(vax_repnum, 2)),
    hoverinfo = "text",
    visible = (r0_val == 10),
    name = paste0("R0 = ", r0_val)
  )
}

steps_rank <- lapply(seq_along(R0_baseline_range), function(i) {
  vis <- rep(FALSE, length(R0_baseline_range))
  vis[i] <- TRUE
  list(method = "restyle", args = list("visible", as.list(vis)), label = as.character(R0_baseline_range[i]))
})

p_rank %>% layout(
  title = "Risk Identification: The Model vs. Raw Vaccination Rates",
  xaxis = list(title = "Rank Difference (Vax Rate Rank \u2212 Model Rv Rank)"),
  yaxis = list(title = ""),
  sliders = list(list(active = 2, currentvalue = list(prefix = "Baseline R0 = "), steps = steps_rank)),
  showlegend = FALSE, margin = list(l = 100)
)
```

3. Threshold Testing Across Transmissibility Scenarios

Pathogen contagiousness can vary. As the baseline R0​ of a strain increases, more counties cross the critical threshold of Rv​≥1, shifting from herd immunity to being actively vulnerable to outbreaks.

```{r rv-sensitivity, cache=TRUE}
vaxrepnum_sensitivity <- data.frame()

for (R0_test in R0_baseline_range) {
  for (county_name in names(utah_counties)) {
    c_short <- gsub(" County", "", county_name)
    popsize <- utah_counties[[county_name]]$age_pops
    initV <- round(county_immunity[[county_name]] * popsize)
    R0_eff <- R0_test * county_R0_factors$R0_factor[county_R0_factors$county == c_short]
    trmat <- transmissionRates(R0_eff, 1, county_contact_matrices[[c_short]])
    vax_Rn <- Re(vaxrepnum(1, popsize, trmat, rep(0, length(popsize)), initV, 1))
    
    vaxrepnum_sensitivity <- rbind(vaxrepnum_sensitivity, data.frame(
      baseline_R0 = R0_test, county = c_short, vax_repnum = vax_Rn
    ))
  }
}

threshold_counts <- vaxrepnum_sensitivity %>%
  group_by(baseline_R0) %>%
  summarise(n_herd_immunity = sum(vax_repnum < 1), n_vulnerable = sum(vax_repnum >= 1), .groups = "drop") %>%
  pivot_longer(cols = c(n_herd_immunity, n_vulnerable), names_to = "status", values_to = "n_counties") %>%
  mutate(status = ifelse(status == "n_herd_immunity", "Herd Immunity Maintained", "Vulnerable to Outbreaks"))

ggplot(threshold_counts, aes(x = baseline_R0, y = n_counties, fill = status)) +
  geom_area(alpha = 0.8) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "white", size=1) +
  scale_fill_manual(values = c("Herd Immunity Maintained" = "#2ca02c", "Vulnerable to Outbreaks" = "#d62728")) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Statewide Vulnerability Under Varying Pathogen Contagiousness",
       x = "Assumed Baseline R0", y = "Number of Counties", fill = "County Status") +
  theme_minimal()
```

Interactive County Vulnerability (Rv​)

Use the slider below to adjust the baseline pathogen contagiousness (R0​) and observe which specific counties cross the Rv​=1 herd immunity threshold.

```{r}
county_pops <- data.frame(county = gsub(" County", "", names(utah_counties)),
                          population = sapply(utah_counties, function(x) sum(x$age_pops)))
rv_plotly_data <- merge(vaxrepnum_sensitivity, county_pops, by = "county") %>%
  arrange(baseline_R0, vax_repnum)

rv_order <- rv_plotly_data %>% filter(baseline_R0 == 10) %>% arrange(vax_repnum) %>% pull(county)
rv_plotly_data$county_f <- factor(rv_plotly_data$county, levels = rv_order)

p <- plot_ly()
for (r0_val in sort(unique(rv_plotly_data$baseline_R0))) {
  df_sub <- rv_plotly_data %>% filter(baseline_R0 == r0_val)
  p <- p %>% add_trace(
    data = df_sub, x = ~vax_repnum, y = ~county_f, type = "bar", orientation = "h",
    marker = list(color = ifelse(df_sub$vax_repnum >= 1, "#d62728", "#2ca02c")),
    text = ~scales::comma(population),
    textposition = "outside",
    hovertext = ~paste0(county, "<br>Rv: ", round(vax_repnum, 2), "<br>Pop: ", scales::comma(population)),
    hoverinfo = "text", visible = (r0_val == 10), name = paste0("R0 = ", r0_val)
  )
}

steps <- lapply(seq_along(unique(rv_plotly_data$baseline_R0)), function(i) {
  vis <- rep(FALSE, length(unique(rv_plotly_data$baseline_R0)))
  vis[i] <- TRUE
  list(method = "restyle", args = list("visible", as.list(vis)), label = as.character(unique(rv_plotly_data$baseline_R0)[i]))
})

p %>% layout(
  title = "Vaccination Reproduction Number (Rv) by County",
  xaxis = list(title = "Vaccination Reproduction Number (Rv)"),
  yaxis = list(title = ""),
  shapes = list(list(type = "line", x0 = 1, x1 = 1, y0 = -0.5, y1 = length(unique(rv_plotly_data$county)) - 0.5, line = list(color = "black", dash = "dash"))),
  sliders = list(list(active = 2, currentvalue = list(prefix = "Baseline R0 = "), steps = steps)),
  showlegend = FALSE, margin = list(l = 100)
)
```

4. Outbreak Consequence (Pure Stochastic Simulation)

If an infection is introduced, Rv​ tells us if it can spread, but pure stochastic simulations tell us how bad it might get.

For this analysis, we run 250 independent simulations per county (assuming introduction by a young adult 18-25).

    Major Outbreak Threshold: We define a major outbreak as >100 absolute cases.

    Herd Immunity Logic: If a county's Rv​<1, the mathematical probability of a major outbreak is forced to 0, as transmission chains will naturally extinguish.

```{r stochastic-sims, cache=TRUE}
nsims_stochastic <- 250

results_list <- vector("list", length(R0_baseline_range) * length(utah_counties))
sim_list     <- vector("list", length(R0_baseline_range) * length(utah_counties))
idx <- 1L

for (R0_stochastic in R0_baseline_range) {
  for (county_name in names(utah_counties)) {
    c_short <- gsub(" County", "", county_name)
    popsize <- utah_counties[[county_name]]$age_pops
    initV <- round(county_immunity[[county_name]] * popsize)
    initI <- rep(0, length(popsize)); initI[6] <- 1  # Intro in young adults
    
    R0_eff <- R0_stochastic * county_R0_factors$R0_factor[county_R0_factors$county == c_short]
    trmat <- transmissionRates(R0_eff, 1, county_contact_matrices[[c_short]])
    vax_Rn <- Re(vaxrepnum(1, popsize, trmat, rep(0, length(popsize)), initV, 1))
    
    # Run pure stochastic simulation
    fs_stochastic <- finalsize(popsize, R0_eff, county_contact_matrices[[c_short]], 
                               rep(1, length(popsize)), rep(1, length(popsize)), 
                               rep(0, length(popsize)), initI, initV, 
                               method = "stochastic", nsims = nsims_stochastic, nthreads = -1)
    
    total_infected <- rowSums(fs_stochastic)
    
    # Logic Gate: If Rv < 1, probability of major escape is effectively 0
    prob_major <- ifelse(vax_Rn < 1, 0, mean(total_infected > 100))
    
    results_list[[idx]] <- data.frame(
      baseline_R0 = R0_stochastic,
      county = c_short,
      population = sum(popsize),
      vax_repnum = vax_Rn,
      prob_major_outbreak = prob_major
    )
    
    # Store all sims for Rv >= 1 counties (bimodal distribution visible)
    if (vax_Rn >= 1) {
      sim_list[[idx]] <- data.frame(
        baseline_R0 = R0_stochastic,
        county = c_short,
        total_cases = total_infected,
        is_major = total_infected > 100
      )
    }
    
    idx <- idx + 1L
  }
}

stochastic_results <- bind_rows(results_list)
sim_long           <- bind_rows(sim_list)
```

Probability of a Major Outbreak (>100 Cases)

This chart shows the real-world consequence probability. Counties missing from this list possess a protective Rv​<1 (herd immunity). Use the slider to adjust the baseline R0​.

```{r}
# Counties that have prob > 0 at any R0 level
all_prob_counties <- stochastic_results %>%
  filter(prob_major_outbreak > 0) %>% pull(county) %>% unique()
prob_order <- stochastic_results %>%
  filter(baseline_R0 == 10, county %in% all_prob_counties) %>%
  arrange(prob_major_outbreak) %>% pull(county)

p_prob <- plot_ly()
for (r0_val in R0_baseline_range) {
  df_sub <- stochastic_results %>%
    filter(baseline_R0 == r0_val, county %in% all_prob_counties)
  df_sub$county_f <- factor(df_sub$county, levels = prob_order)

  p_prob <- p_prob %>% add_trace(
    data = df_sub, x = ~prob_major_outbreak * 100, y = ~county_f,
    type = "bar", orientation = "h",
    marker = list(color = "#8b08b0"),
    text = ~paste0(round(prob_major_outbreak * 100, 1), "%"),
    textposition = "outside",
    hovertext = ~paste0(county, "<br>Prob: ", round(prob_major_outbreak * 100, 1), "%"),
    hoverinfo = "text",
    visible = (r0_val == 10),
    name = paste0("R0 = ", r0_val)
  )
}

steps_prob <- lapply(seq_along(R0_baseline_range), function(i) {
  vis <- rep(FALSE, length(R0_baseline_range))
  vis[i] <- TRUE
  list(method = "restyle", args = list("visible", as.list(vis)), label = as.character(R0_baseline_range[i]))
})

p_prob %>% layout(
  title = "Probability of Outbreak Exceeding 100 Cases",
  xaxis = list(title = "Probability (%)", range = c(0, 105)),
  yaxis = list(title = ""),
  sliders = list(list(active = 2, currentvalue = list(prefix = "Baseline R0 = "), steps = steps_prob)),
  showlegend = FALSE, margin = list(l = 100)
)
```

Full Outbreak Distribution

All 250 simulations per county are shown (for counties with Rv​≥1). The bimodal distribution — a spike at zero (chains that fizzle out) and a heavier right tail (sustained outbreaks) — is a hallmark of stochastic transmission near the herd immunity threshold.

```{r}
if(nrow(sim_long) > 0) {
  # Order by median outbreak size at R0 = 10
  order_all <- sim_long %>%
    filter(baseline_R0 == 10) %>%
    group_by(county) %>% summarise(med = median(total_cases)) %>% arrange(med) %>% pull(county)
  all_sim_counties <- unique(sim_long$county)
  order_all_full <- c(order_all, sort(setdiff(all_sim_counties, order_all)))

  p_dots <- plot_ly()
  for (r0_val in R0_baseline_range) {
    df_sub <- sim_long %>% filter(baseline_R0 == r0_val)
    df_sub$county_f <- factor(df_sub$county, levels = order_all_full)
    # Color: red = major outbreak (>100), grey = fizzled
    dot_colors <- ifelse(df_sub$is_major, "#d62728", "#aaaaaa")

    p_dots <- p_dots %>% add_trace(
      data = df_sub, x = ~total_cases, y = ~county_f, type = "scatter", mode = "markers",
      marker = list(color = dot_colors, size = 4, opacity = 0.25),
      hovertext = ~paste0(county, "<br>Cases: ", scales::comma(total_cases),
                          "<br>", ifelse(is_major, "Major outbreak", "Fizzled")),
      hoverinfo = "text",
      visible = (r0_val == 10),
      name = paste0("R0 = ", r0_val)
    )
  }

  steps_dots <- lapply(seq_along(R0_baseline_range), function(i) {
    vis <- rep(FALSE, length(R0_baseline_range))
    vis[i] <- TRUE
    list(method = "restyle", args = list("visible", as.list(vis)), label = as.character(R0_baseline_range[i]))
  })

  p_dots %>% layout(
    title = "Full Simulation Distribution of Outbreak Size",
    xaxis = list(title = "Total Cases per Simulation (log scale)", type = "log"),
    yaxis = list(title = ""),
    sliders = list(list(active = 2, currentvalue = list(prefix = "Baseline R0 = "), steps = steps_dots)),
    showlegend = FALSE, margin = list(l = 100)
  )
} else {
  print("No vulnerable counties (Rv >= 1) under current parameters.")
}
```

## 5. Final Risk Summary Matrix {.tabset}

For each county, the **Risk Score** is defined as:

$$\text{Risk Score} = \frac{P_i(x > 100)}{P_t} \times \text{Population}_i$$

where $P_i(x > 100)$ is the probability of exceeding 100 cases in county $i$ and $P_t$ is the total statewide population. This quantity represents the expected number of at-risk individuals contributed by each county relative to the state as a whole.

```{r}
total_pop <- sum(unique(stochastic_results[, c("county", "population")])$population)

make_risk_table <- function(r0_val) {
  stochastic_results %>%
    filter(baseline_R0 == r0_val) %>%
    mutate(
      risk_score = (prob_major_outbreak * population) / total_pop,
      `Rv` = round(vax_repnum, 3),
      `Prob. >100 Cases (%)` = round(prob_major_outbreak * 100, 1),
      `Risk Score` = round(risk_score, 4),
      `Population` = scales::comma(population)
    ) %>%
    select(County = county, `Population`, `Rv`, `Prob. >100 Cases (%)`, `Risk Score`) %>%
    arrange(desc(`Risk Score`))
}
```

### R0 = 6

```{r}
knitr::kable(make_risk_table(6), caption = "Risk Matrix (Baseline R\u2080 = 6)")
```

### R0 = 8

```{r}
knitr::kable(make_risk_table(8), caption = "Risk Matrix (Baseline R\u2080 = 8)")
```

### R0 = 10

```{r}
knitr::kable(make_risk_table(10), caption = "Risk Matrix (Baseline R\u2080 = 10)")
```

### R0 = 12

```{r}
knitr::kable(make_risk_table(12), caption = "Risk Matrix (Baseline R\u2080 = 12)")
```

### R0 = 14

```{r}
knitr::kable(make_risk_table(14), caption = "Risk Matrix (Baseline R\u2080 = 14)")
```
