---
title: "Utah County Risk Score Formulation"
date: "April 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Utah County Risk Score Formulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in any specific location.
DATA: The vaccination data (county level) is not public and so is not included in the package.

This notebook loads precomputed stochastic simulation outputs and focuses only on risk score formulation.

## 1. Load Precomputed Simulation Bundle

```{r load-data, cache=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

bundle_candidates <- c(
  file.path("outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "..", "..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData")
)

existing_bundle <- bundle_candidates[file.exists(bundle_candidates)]
if (length(existing_bundle) == 0) {
  stop("Could not find utah_county_simulation_bundle.RData. Run utah_county_simulations.Rmd first.")
}

bundle_path <- existing_bundle[1]
load(bundle_path)

required_objects <- c(
  "stochastic_results",
  "simulation_stats_by_county",
  "simulation_stats_statewide",
  "R0_baseline",
  "R0_baseline_range",
  "nsims_stochastic"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects) > 0) {
  stop(paste("Missing required objects in .RData:", paste(missing_objects, collapse = ", ")))
}

cat("Loaded:", normalizePath(bundle_path), "\n")
cat("Rows in stochastic_results:", nrow(stochastic_results), "\n")
```

## 2. Internal Scoring Design

At this stage, outputs are internal and county-resolved. We still use a relative formulation so scores communicate comparative risk, not deterministic forecasts.

### Decision Guide (Briefing Use)

- Use `risk_relative_mean1` for county-level interpretation within a given baseline $R_0$ scenario.
- Use `risk_latent` trends across $R_0$ values to understand how risk structure shifts as transmissibility assumptions change.
- Treat scores as comparative indicators, not forecasts of exact outbreak counts.
- Use component contributions (`burden_weighted`, `major_weighted`, `transmission_weighted`, `tail_weighted`) to explain *why* a county is elevated.

### Assumptions and Limitations

- Weights are heuristic (`0.35/0.35/0.20/0.10`) and encode policy judgment about component importance.
- `percentile01` is rank-based: it preserves ordering, not absolute distance between counties.
- Missing component values are replaced with `0` before percentile ranking.
- Relative scores are scenario-specific; they should not be interpreted as absolute risk probabilities.
- Results depend on simulation design choices (seed distribution, $k$ threshold for major outbreaks, and modeled vaccine effects).

## 3. Relative Risk Score Formulation

For county $c$ in baseline scenario $r$, define four normalized components:

$$
\begin{aligned}
B_{c,r} &= \text{percentile01}\left(\log(1 + \text{expected cases per 10k})\right) \\
M_{c,r} &= \text{percentile01}\left(P(\text{outbreak} > k)\right) \\
E_{c,r} &= \text{percentile01}\left(\max(R_v - 1, 0)\right) \\
T_{c,r} &= \text{percentile01}\left(\log(1 + q_{95}(\text{cases per 10k}))\right)
\end{aligned}
$$

`percentile01` converts each county's raw metric into a simple relative position from 0 to 1, compared only against other counties in the same baseline $R_0$ scenario.

1. Take one metric (for example, major-outbreak probability) and sort counties from lowest to highest.
2. Convert each county's position in that sorted list to a 0-1 scale.
3. Use that 0-1 value in the weighted score.

So for a given component, a county near `0` is near the low end among peers, a county near `1` is near the high end, and `0.5` is around the middle.

Mathematically, for component value $x_{c,r}$:

$$
\operatorname{percentile01}(x_{c,r}) = \frac{\operatorname{rank}_r(\widetilde{x}_{c,r}) - 1}{N_r - 1},
\qquad
\widetilde{x}_{c,r} =
\begin{cases}
x_{c,r}, & x_{c,r}\neq \mathrm{NA} \\
0, & x_{c,r}=\mathrm{NA}
\end{cases}
$$

with $N_r$ counties in scenario $r$. This maps values to $[0,1]$ (lowest $\to 0$, highest $\to 1$), and tied values receive the same percentile rank.

With weights $w_B=0.35$, $w_M=0.35$, $w_E=0.20$, $w_T=0.10$:

$$
S_{c,r} = w_B B_{c,r} + w_M M_{c,r} + w_E E_{c,r} + w_T T_{c,r}
$$

The primary internal metric is normalized so county mean risk is 1.0 within each baseline $R_0$ scenario:

$$
\operatorname{RiskRelative}_{c,r} = \frac{S_{c,r}}{\mu_r}
$$

So a value of 1.2 means 20% higher modeled risk than the scenario mean, and 0.8 means 20% lower.

For standardized reporting, we also compute

$$
\operatorname{RiskZ}_{c,r} = \frac{S_{c,r} - \mu_r}{\sigma_r}
$$

where $\mu_r,\sigma_r$ are county-level mean and SD of $S_{c,r}$ within scenario $r$.

### Worked Example (One County)

Suppose one county has component percentiles:

- $B_{c,r}=0.70$, $M_{c,r}=0.60$, $E_{c,r}=0.20$, $T_{c,r}=0.50$

Then:

$$
S_{c,r} = 0.35(0.70) + 0.35(0.60) + 0.20(0.20) + 0.10(0.50) = 0.545
$$

If the scenario mean latent score is $\mu_r=0.50$, then

$$
\operatorname{RiskRelative}_{c,r} = \frac{0.545}{0.50} = 1.09
$$

Interpretation: this county is about 9% above the scenario-average modeled risk level.

### Why These Four Weighted Variables

- **Burden component ($w_B=0.35$)**: captures expected impact size while normalizing by population; answers "how much total burden might this county see relative to peers?"
- **Major outbreak probability ($w_M=0.35$)**: captures event likelihood; answers "how likely is a non-trivial outbreak to occur at all?"
- **Escape/transmission component ($w_E=0.20$)**: uses $\max(R_v - 1, 0)$ to represent escape risk beyond herd-immunity threshold; answers "is sustained transmission still plausible?"
- **Tail severity component ($w_T=0.10$)**: uses the upper-tail quantile ($q_{95}$) to retain rare-but-severe behavior; answers "how bad can outcomes get in adverse realizations?"

The weighting prioritizes expected burden and major-outbreak likelihood, while still retaining mechanistic escape risk and severe-tail information.

```{r risk-formulation}
# Join simulation summary and tail statistics
risk_input <- stochastic_results %>%
  left_join(
    simulation_stats_by_county %>%
      select(baseline_R0, county, p95_cases),
    by = c("baseline_R0", "county")
  ) %>%
  mutate(
    expected_cases_per_10k = ifelse(population > 0, expected_cases / population * 10000, 0),
    tail_cases_per_10k = ifelse(population > 0, p95_cases / population * 10000, 0),
    rv_excess = pmax(vax_repnum - 1, 0)
  )

# Percentile-rank scaling helper (robust to scale and outliers)
percentile01 <- function(x) {
  dplyr::percent_rank(replace_na(x, 0))
}

# Weight set (sum should be 1)
weights <- c(
  burden = 0.35,
  major = 0.35,
  transmission = 0.20,
  tail = 0.10
)

if (abs(sum(weights) - 1) > 1e-9) {
  stop("Risk weights must sum to 1.")
}

risk_scores_internal <- risk_input %>%
  group_by(baseline_R0) %>%
  mutate(
    burden_component = percentile01(log1p(expected_cases_per_10k)),
    major_component = percentile01(prob_major_outbreak),
    transmission_component = percentile01(rv_excess),
    tail_component = percentile01(log1p(tail_cases_per_10k)),
    risk_latent =
      weights[["burden"]] * burden_component +
      weights[["major"]] * major_component +
      weights[["transmission"]] * transmission_component +
      weights[["tail"]] * tail_component,
    latent_center = mean(risk_latent, na.rm = TRUE),
    latent_scale = sd(risk_latent, na.rm = TRUE),
    risk_relative_mean1 = ifelse(latent_center > 0, risk_latent / latent_center, 1),
    risk_relative_latent = risk_relative_mean1,
    risk_relative_z = ifelse(latent_scale > 0, (risk_latent - latent_center) / latent_scale, 0),
    burden_weighted = weights[["burden"]] * burden_component,
    major_weighted = weights[["major"]] * major_component,
    transmission_weighted = weights[["transmission"]] * transmission_component,
    tail_weighted = weights[["tail"]] * tail_component
  ) %>%
  ungroup()
```

## 4. Internal Per-County Breakdown (Restricted)

This table is intended for internal analysis only and includes county-level metrics and each weighted score component.

```{r county-breakdown-internal}
risk_county_breakdown_baseline <- risk_scores_internal %>%
  filter(baseline_R0 == R0_baseline) %>%
  arrange(desc(risk_relative_mean1)) %>%
  transmute(
    county,
    population,
    expected_cases,
    expected_cases_per_10k,
    prob_major_outbreak,
    vax_repnum,
    p95_cases,
    tail_cases_per_10k,
    burden_component,
    major_component,
    transmission_component,
    tail_component,
    burden_weighted,
    major_weighted,
    transmission_weighted,
    tail_weighted,
    risk_latent,
    risk_relative_latent,
    risk_relative_mean1,
    risk_relative_z
  )

knitr::kable(
  risk_county_breakdown_baseline %>%
    mutate(
      population = comma(population),
      expected_cases = round(expected_cases, 2),
      expected_cases_per_10k = round(expected_cases_per_10k, 3),
      prob_major_outbreak = round(prob_major_outbreak, 4),
      vax_repnum = round(vax_repnum, 4),
      p95_cases = round(p95_cases, 1),
      tail_cases_per_10k = round(tail_cases_per_10k, 3),
      burden_component = round(burden_component, 3),
      major_component = round(major_component, 3),
      transmission_component = round(transmission_component, 3),
      tail_component = round(tail_component, 3),
      burden_weighted = round(burden_weighted, 3),
      major_weighted = round(major_weighted, 3),
      transmission_weighted = round(transmission_weighted, 3),
      tail_weighted = round(tail_weighted, 3),
      risk_latent = round(risk_latent, 3),
      risk_relative_latent = round(risk_relative_latent, 3),
      risk_relative_mean1 = round(risk_relative_mean1, 3),
      risk_relative_z = round(risk_relative_z, 3)
    ),
  caption = paste0("Internal county-level metric breakdown for baseline R0 = ", R0_baseline)
)
```

## 5. Risk Score Definitions

The model reports several risk score variants for complementary purposes:

- `risk_latent`: the primary weighted composite score, defined as
  $$
  S_{c,r} = w_B B_{c,r} + w_M M_{c,r} + w_E E_{c,r} + w_T T_{c,r}
  $$
  where larger values indicate higher modeled relative risk burden for that county/scenario.
- `risk_relative_latent` and `risk_relative_mean1`: mean-normalized latent score within each baseline $R_0$,
  $$
  \frac{S_{c,r}}{\mu_r}
  $$
  so `1.0` is the scenario mean, `1.2` is 20% above mean, and `0.8` is 20% below mean.
- `risk_relative_z`: standardized latent score,
  $$
  \frac{S_{c,r} - \mu_r}{\sigma_r}
  $$
  interpreted in standard deviation units.

For cross-$R_0$ trend visualization, `risk_latent` is preferred. For within-scenario county interpretation, `risk_relative_mean1` is usually easiest to communicate.

## 6. County Risk Scores Only

```{r county-scores-only}
risk_scores_only_baseline <- risk_scores_internal %>%
  filter(baseline_R0 == R0_baseline) %>%
  arrange(desc(risk_relative_mean1)) %>%
  transmute(
    county,
    risk_relative_latent,
    risk_relative_z
  )

knitr::kable(
  risk_scores_only_baseline %>%
    mutate(
      risk_relative_latent = round(risk_relative_latent, 3),
      risk_relative_z = round(risk_relative_z, 3)
    ),
  caption = paste0("County risk scores only for baseline R0 = ", R0_baseline)
)
```

## 7. Latent Risk Across R0 Scenarios

```{r risk-across-r0-internal}
risk_latent_summary <- risk_scores_internal %>%
  group_by(baseline_R0) %>%
  summarise(
    mean_latent = mean(risk_latent, na.rm = TRUE),
    p25_latent = quantile(risk_latent, 0.25, na.rm = TRUE),
    p50_latent = quantile(risk_latent, 0.50, na.rm = TRUE),
    p75_latent = quantile(risk_latent, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(risk_latent_summary, aes(x = baseline_R0, y = p50_latent)) +
  geom_ribbon(aes(ymin = p25_latent, ymax = p75_latent), alpha = 0.25, fill = "#3182bd") +
  geom_line(linewidth = 1.0, color = "#08519c") +
  geom_line(aes(y = mean_latent), linewidth = 0.9, linetype = "dashed", color = "#525252") +
  labs(
    title = "Latent Risk Across Baseline R0",
    x = "Baseline R0",
    y = "Latent risk score"
  ) +
  theme_minimal()
```

## 8. Export Outputs

```{r export-risk-scores, cache=FALSE}
export_dir <- file.path("outputs", "utah_county_risk")
if (!dir.exists(export_dir)) dir.create(export_dir, recursive = TRUE)

# Internal all-scenario scores
write.csv(
  risk_scores_internal,
  file.path(export_dir, "risk_scores_internal_all_r0.csv"),
  row.names = FALSE
)

# Internal county-level breakdown for baseline scenario
write.csv(
  risk_county_breakdown_baseline,
  file.path(export_dir, "risk_scores_internal_county_breakdown_baseline_r0.csv"),
  row.names = FALSE
)

write.csv(
  risk_scores_only_baseline,
  file.path(export_dir, "risk_scores_internal_only_baseline_r0.csv"),
  row.names = FALSE
)

write.csv(
  risk_latent_summary,
  file.path(export_dir, "risk_scores_internal_latent_summary_by_r0.csv"),
  row.names = FALSE
)

# Internal (contains county names): keep private and do not publish
save(
  risk_scores_internal,
  risk_county_breakdown_baseline,
  risk_scores_only_baseline,
  risk_latent_summary,
  weights,
  file = file.path(export_dir, "risk_scores_internal_private.RData")
)

cat("Saved risk score outputs to:", normalizePath(export_dir), "\n")
cat("Outputs are internal-only and include county-level breakdowns.\n")
```
