---
title: "Measles Age-Structured Model"
output: rmarkdown::md_document
vignette: >
  %\VignetteIndexEntry{Measles Age-Structured Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7
)
```

```{r setup}
library(multigroup.vaccine)
library(socialmixr)
library(ggplot2)

verbose <- FALSE  # Set to TRUE to enable detailed logging during processing
```

# Age-Structured UT County Model

This vignette is an experimental vignette in which we build and run an age-structured model of transmission within Utah on a county-level resolution.

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in this location.

## Age Structures

First, define the age groups for our model.

Because each county has a different school vax distribution, we need to use the same age groups as the school vax distribution for a given county.
In the vaccination data (county_vax_data.csv), the vaccination data is organized by **school** level (1=el, 2=middle, 3=high).

**Important methodological note on grade structure assumptions:**

School grade structures in Utah are determined at the **district level**, not county level. Within a single county, different districts may use different configurations (e.g., some K-5/6-8/9-12, others K-6/7-9/10-12). However, our vaccination data is aggregated at the **county level**.

For modeling purposes, we make the following **simplifying assumptions**:

1. **County-level aggregation is acceptable**: Since our vaccination data is county-aggregated, we treat each county as having a *dominant grade structure*
2. **Population ratios reveal dominant patterns**: We use the relative sizes of elementary/middle/high school populations to infer the most common grade configuration in that county
3. **Age groups approximate school levels**: Our age groups are designed to roughly align with school levels, acknowledging that exact grade-to-age mapping varies by district policies

This approach works best for counties with relatively uniform district structures.
Counties with highly heterogeneous district structures may require manual overrides based on demographic research.

### Heuristics

For the first (somewhat naive approach), we will use the *ratio* of the school populations to determine the age groups.

**Inference heuristic for age groups based on population ratios:**

The logic uses population ratios to infer grade structures:
- If middle ≈ 50% of high: Middle likely 2 grades (7-8) → elementary = K-6
- If middle ≈ 75% of high: Middle likely 3 grades (6-8) → elementary = K-5
- If middle ≈ high (ratio 80%-120%):
  - if elementary is much smaller than middle (ratio <0.5): likely K-4 for elementary, which supports 4-grade structures for middle and high (5-8, 9-12)
  - if elementary is larger than middle (ratio >0.5): likely K-6 for elementary, which supports 3-grade structures for middle and high (7-9, 10-12) (ie: Weber County fits this pattern)

**Limitations of this heuristic:**
- Assumes uniform grade structure across all districts within a county
- Does not account for charter schools or alternative school configurations
- Students enrolled in *online* schools may not be captured accurately in the population ratios
- Population ratios may be influenced by factors other than grade spans (e.g., demographic patterns, school choice)
- Manual overrides can be added for counties where specific district information is known (for example, Millard County has a known K-4/5-8/9-12 structure based on district data)

**Manual overrides:** For counties with verified grade structures that differ from the heuristic inference, add them to the `manual_overrides` list in the code below.

```{r define-age-groups}
county_vax_data <- read.csv(file.path("../../data-raw/county_vax_data.csv"))

age_groups_by_county <- list()

# Get unique counties
unique_counties <- unique(county_vax_data$county)

for (county_name in unique_counties) {
  # Extract data for this county
  county_data <- county_vax_data[county_vax_data$county == county_name, ]
  
  # Get population for each level
  el_pop <- county_data$pop[county_data$level == 1]
  middle_pop <- county_data$pop[county_data$level == 2]
  high_pop <- county_data$pop[county_data$level == 3]
  
  if (verbose) {
    print(paste("Processing county:", county_name))
    print(paste("Elementary pop:", el_pop, "Middle pop:", middle_pop, "High pop:", high_pop))
  }
  
  # Some counties have more unique structures, so we can define manual overrides for those based on known district configurations
  # This is especially important for rural counties.
  manual_overrides <- list(
    # Example:
    # "Millard" = list(
    #   ages = c(0, 5, 10, 14, 18, 25, 45, 70),
    #   reason = "K-4 elementary, 5-8 middle, 9-12 high (verified from district data)"
    # )
  )

  # Check for manual override first
  if (county_name %in% names(manual_overrides)) {
    override <- manual_overrides[[county_name]]
    age_groups_by_county[[county_name]] <- override$ages
    if (verbose) {
      print(paste("  -> Using manual override:", override$reason))
      print(paste("     Ages:", paste(override$ages, collapse=", ")))
    }
  } else if (length(middle_pop) > 0 && length(high_pop) > 0 && high_pop > 0) {
    # Use population ratio to determine age structure
    middle_to_high_ratio <- middle_pop / high_pop
    el_to_middle_ratio <- if(length(el_pop) > 0 && middle_pop > 0) el_pop / middle_pop else NA

    if (verbose) {
      print(paste("Middle to high ratio:", round(middle_to_high_ratio, 3)))
      if (!is.na(el_to_middle_ratio)) {
        print(paste("Elementary to middle ratio:", round(el_to_middle_ratio, 3)))
      }
    }

    # logic to detect different grade structures
    if (middle_to_high_ratio >= 0.8 && middle_to_high_ratio <= 1.2) {  # 1.2 is used to provide 'padding' for small population noise
      # if the elementary to middle ratio is low (e.g., <1.3), it may indicate a K-4 structure for elementary, which would further support the 4-grade inference for middle and high
      # however, if the elementary to middle ratio is higher (e.g., >1.3), it may indicate a K-6 structure for elementary, which would suggest a 3-grade structure for middle (7-9) and 3-grade for high (10-12)
      if (!is.na(el_to_middle_ratio) && el_to_middle_ratio < 1.3) {
        # Elementary is much smaller than middle → likely K-4
        # Suggests: Elementary K-4, Middle 5-8, High 9-12
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 10, 14, 18, 25, 45, 70)
        if (verbose) print("  -> Using ages: 0, 1, 5, 10, 14, 18, 25, 45, 70 (inferred: el=K-4, middle=5-8, high=9-12)")
      } else {
        # Elementary is larger than middle → likely K-6
        # Suggests: Elementary K-5, Middle 7-9, High 10-12
        age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 15, 18, 25, 45, 70)
        if (verbose) print("  -> Using ages: 0, 1, 5, 12, 15, 18, 25, 45, 70 (inferred: el=K-5, middle=7-9, high=10-12)")
      }
    } else if (middle_to_high_ratio <= 0.6) {
      # Middle is ~50% of high → middle likely 2 grades (7-8)
      # Suggests: Elementary K-6, Middle 7-8, High 9-12
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 12, 14, 18, 25, 45, 70)
      if (verbose) print("  -> Using ages: 0, 1, 5, 12, 14, 18, 25, 45, 70 (inferred: el=K-6, middle=7-8, high=9-12)")
    } else {
      # Middle is ~60-80% of high → middle likely 3 grades (6-8)
      # Suggests: Elementary K-5, Middle 6-8, High 9-12
      age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
      if (verbose) print("  -> Using ages: 0, 1, 5, 11, 14, 18, 25, 45, 70 (inferred: el=K-5, middle=6-8, high=9-12)")
    }
  } else {
    # Default if data is missing
    age_groups_by_county[[county_name]] <- c(0, 1, 5, 11, 14, 18, 25, 45, 70)
    if (verbose) print("  -> Using default ages: 0, 1, 5, 11, 14, 18, 25, 45, 70")
  }
  if (verbose) print("")
}

if (verbose) {
  print("Age groups by county:")
  print(age_groups_by_county)
  
  print("Number of counties processed:")
  print(length(age_groups_by_county))
}

# One of the 'counties' is Online, which represents online schools in UT.
# Remove that county from the list and save it as a separate variable
online_schools_age_groups <- age_groups_by_county$Online
age_groups_by_county$Online <- NULL

if (verbose) {
  print("Age groups for online schools:")
  print(online_schools_age_groups)
}
```

Now get all counties from Utah, passing in the age groups for each county as defined above:

```{r get-all-counties}
# get all counties from Utah using getAllCountiesData with county-specific age groups
# data is pre-saved as csv in data-raw
utah_counties <- getAllCountiesData(
  state_fips = getStateFIPS("Utah"),
  year = 2024,
  age_groups_by_county = age_groups_by_county,
  csv_path = getCensusDataPath(),
  verbose = verbose
)

# Display the names of all counties
names(utah_counties)

# Show example data from one county
if (verbose) {
  print("Example county data:")
  print(utah_counties[[1]])
}
```

```{r visualize-age-structures}
# Categorize each county by its age group structure
# Extract the school-age portions (indices 3-5) to identify the structure type
structure_df <- data.frame(
  county = character(),
  structure_type = character(),
  el_start = numeric(),
  el_end = numeric(),
  middle_start = numeric(),
  middle_end = numeric(),
  high_start = numeric(),
  high_end = numeric(),
  stringsAsFactors = FALSE
)

for (county_name in names(age_groups_by_county)) {
  ages <- age_groups_by_county[[county_name]]

  # School age groups are typically at indices 3, 4, 5, 6
  # (after 0, 1, and the first school boundary)
  # For visualization, we want segments to connect, so end points should equal the next start point
  el_start <- ages[3]
  el_end <- ages[4]  # This connects to middle_start
  middle_start <- ages[4]
  middle_end <- ages[5]  # This connects to high_start
  high_start <- ages[5]
  high_end <- ages[6]  # This is the final boundary

  # Categorize based on the age structure
  if (identical(ages, c(0, 1, 5, 10, 14, 18, 25, 45, 70))) {
    structure_type <- "K-4 / 5-8 / 9-12"
  } else if (identical(ages, c(0, 1, 5, 12, 15, 18, 25, 45, 70))) {
    structure_type <- "K-6 / 7-9 / 10-12"
  } else if (identical(ages, c(0, 1, 5, 12, 14, 18, 25, 45, 70))) {
    structure_type <- "K-6 / 7-8 / 9-12"
  } else if (identical(ages, c(0, 1, 5, 11, 14, 18, 25, 45, 70))) {
    structure_type <- "K-5 / 6-8 / 9-12"
  } else {
    structure_type <- "Other"
  }

  structure_df <- rbind(structure_df, data.frame(
    county = gsub(" County", "", county_name),
    structure_type = structure_type,
    el_start = el_start,
    el_end = el_end,
    middle_start = middle_start,
    middle_end = middle_end,
    high_start = high_start,
    high_end = high_end
  ))
}

# Count by structure type
structure_counts <- table(structure_df$structure_type)
print("Distribution of grade structures:")
print(structure_counts)

# Create a visual showing the grade spans for each county
library(tidyr)
library(dplyr)

# Reshape data for plotting - create one row per county per school level
structure_long <- rbind(
  structure_df %>% select(county, structure_type, start = el_start, end = el_end) %>% mutate(level = "Elementary"),
  structure_df %>% select(county, structure_type, start = middle_start, end = middle_end) %>% mutate(level = "Middle"),
  structure_df %>% select(county, structure_type, start = high_start, end = high_end) %>% mutate(level = "High")
)

# Order counties by structure type and then alphabetically
structure_long <- structure_long %>%
  arrange(structure_type, county) %>%
  mutate(county = factor(county, levels = unique(county)))

# Plot showing grade spans for each county
ggplot(structure_long, aes(y = county)) +
  geom_segment(aes(x = start, xend = end, yend = county, color = level), 
               size = 6, alpha = 0.8) +
  scale_color_manual(values = c("Elementary" = "#1f77b4", 
                                  "Middle" = "#ff7f0e", 
                                  "High" = "#2ca02c")) +
  scale_x_continuous(breaks = seq(0, 20, by = 1), expand = c(0, 0)) +
  facet_grid(structure_type ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Grade Structure Age Spans by County",
    subtitle = "Counties grouped by inferred grade configuration",
    x = "Age",
    y = "County",
    color = "School Level"
  ) +
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0, hjust = 0),
    legend.position = "bottom"
  )
```

### Difference between Vax and Census data

There is a non-negligible difference between the census 2024 estimate age-grouping population sizes and the population sizes from the vaccination data.

This could be due to a variety of external factors.

To understand this difference, we can compare the processed age-groupings and the population sizes from the vaccination data for every county.

```{r compare-vax-census}
county_difference <- data.frame(
  county = character(),
  census_population = numeric(),
  vax_population = numeric(),
  difference = numeric(),
  stringsAsFactors = FALSE
)

# For every county, compute the difference at each age group between the census population and the vaccination data population
for (county_name in names(utah_counties)) {
  census_data <- utah_counties[[county_name]]

  # Remove " County" from the county name to match the format in the vaccination data
  county_name <- gsub(" County", "", county_name)
  vax_data <- county_vax_data[county_vax_data$county == county_name, ]

  if (verbose) print(paste("County:", county_name))

  # Get the population sizes for the school age groups from the census data and the vaccination data
  census_age_1 <- census_data$age_labels[3]  # age group corresponding to the first school age group (e.g., 1-4 or 1-5)
  census_age_2 <- census_data$age_labels[4]  # age group corresponding to the second school age group (e.g., 5-11 or 6-11)
  census_age_3 <- census_data$age_labels[5]  # age group corresponding to the third school age group (e.g., 12-17)

  census_pop <- census_data$age_pops[census_data$age_labels %in% c(census_age_1, census_age_2, census_age_3)]

  vax_pop <- c(vax_data$pop[vax_data$level == 1], vax_data$pop[vax_data$level == 2], vax_data$pop[vax_data$level == 3])

  difference <- sum(census_pop) - sum(vax_pop)

  county_difference <- rbind(county_difference, data.frame(
    county = county_name,
    census_population = sum(census_pop),
    vax_population = sum(vax_pop),
    difference = difference
  ))

  if (verbose) {
    print(paste("County:", county_name))
    print(paste("Census population:", sum(census_pop), "Vax population:",
                sum(vax_pop), "Difference:", sum(census_pop) - sum(vax_pop)))
    print("")
  }
}
```

```{r visualize-age-differences}
print(county_difference)

county_difference$total_population <- county_difference$census_population
county_difference$percent_difference <- abs((county_difference$difference / county_difference$census_population) * 100)

# Bar chart: absolute difference in population size between census and vaccination data for each county
ggplot(county_difference, aes(x = reorder(county, difference), y = difference)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(round(percent_difference, 1), "%")), vjust = -0.5, size = 3) +
  labs(
    title = "Difference in Population Size between Census and Vaccination Data by County",
    subtitle = "Labels show percentage difference",
    x = "County",
    y = "Difference in Population Size"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Two-panel view showing both absolute and percentage differences
library(gridExtra)
p1 <- ggplot(county_difference, aes(x = total_population, y = difference)) +
  geom_point(color = "darkblue", size = 3, alpha = 0.7) +
  geom_text(aes(label = county), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Difference",
    x = "Census School-Age Population (log scale)",
    y = "Difference (Census - Vax)"
  ) +
  theme_minimal()

p2 <- ggplot(county_difference, aes(x = total_population, y = percent_difference)) +
  geom_point(color = "darkred", size = 3, alpha = 0.7) +
  geom_text(aes(label = county), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Percentage Difference",
    x = "Census School-Age Population (log scale)",
    y = "Percentage Difference"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 1)
```