---
title: "Utah County Risk Score Formulation"
date: "April 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Utah County Risk Score Formulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

WARNING: The data and model used in this vignette are for demonstration purposes only and do not reflect real-world conditions accurately. The results are not intended to forecast or predict a measles outbreak size in any specific location.
DATA: The vaccination data (county level) is not public and so is not included in the package.

This notebook loads precomputed stochastic simulation outputs and focuses only on risk score formulation.

## 1. Load Precomputed Simulation Bundle

```{r load-data, cache=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

bundle_candidates <- c(
  file.path("outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData"),
  file.path("..", "..", "..", "outputs", "utah_county_risk", "utah_county_simulation_bundle.RData")
)

existing_bundle <- bundle_candidates[file.exists(bundle_candidates)]
if (length(existing_bundle) == 0) {
  stop("Could not find utah_county_simulation_bundle.RData. Run utah_county_simulations.Rmd first.")
}

bundle_path <- existing_bundle[1]
load(bundle_path)

if (!exists("county_static_metrics")) {
  static_metric_candidates <- c(
    file.path("outputs", "utah_county_risk", "county_static_metrics.csv"),
    file.path("..", "outputs", "utah_county_risk", "county_static_metrics.csv"),
    file.path("..", "..", "outputs", "utah_county_risk", "county_static_metrics.csv"),
    file.path("..", "..", "..", "outputs", "utah_county_risk", "county_static_metrics.csv")
  )

  existing_static_metrics <- static_metric_candidates[file.exists(static_metric_candidates)]
  if (length(existing_static_metrics) > 0) {
    county_static_metrics <- read.csv(existing_static_metrics[1], stringsAsFactors = FALSE)
  }
}

if (!exists("large_outbreak_threshold_frac")) {
  large_outbreak_threshold_frac <- 0.01
}

if (!exists("large_outbreak_min_cases")) {
  large_outbreak_min_cases <- 20
}

if (!"prob_large_outbreak" %in% names(stochastic_results) &&
    "prob_major_outbreak" %in% names(stochastic_results)) {
  stochastic_results$prob_large_outbreak <- stochastic_results$prob_major_outbreak
}

if (!"large_outbreak_threshold_cases" %in% names(stochastic_results) &&
    "susceptible_population" %in% names(stochastic_results)) {
  stochastic_results$large_outbreak_threshold_cases <- stochastic_results$susceptible_population * large_outbreak_threshold_frac
}

required_objects <- c(
  "stochastic_results",
  "simulation_stats_by_county",
  "simulation_stats_statewide",
  "county_static_metrics",
  "R0_baseline",
  "R0_baseline_range",
  "nsims_stochastic"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects) > 0) {
  stop(paste("Missing required objects in .RData:", paste(missing_objects, collapse = ", ")))
}

cat("Loaded:", normalizePath(bundle_path), "\n")
cat("Rows in stochastic_results:", nrow(stochastic_results), "\n")
```

## 2. Tunable Scoring Design

This notebook now emphasizes directly interpretable outputs first and uses simple,
tunable score formulas only as an optional summary layer. The main goal is to
avoid hiding epidemiologically meaningful quantities behind percentile ranks.

### Decision Guide

- Use the raw metrics in the exported CSV whenever possible: `school_vax_rate`, `county_immunity_pct`, `county_susceptible_pct`, `vax_repnum`, `prob_large_outbreak`, and `expected_cases_per_10k`.
- Use `vulnerability_score` to answer "how favorable are conditions for sustained transmission here?"
- Use `priority_score` to answer "where would an outbreak create the most operational burden?"

### Tunable Parameters

The composite scores are defined with explicit caps and weights so the influence
of each raw metric can be adjusted transparently.

```{r score-parameters}
score_params <- list(
  large_outbreak_threshold_frac = large_outbreak_threshold_frac,
  large_outbreak_min_cases = large_outbreak_min_cases,
  rv_excess_cap = 1.0,
  susceptible_pct_cap = 0.15,
  per_10k_population_floor = 10000,
  expected_cases_per_10k_cap = 100,
  tail_cases_per_10k_cap = 250,
  expected_cases_cap = 250,
  p95_cases_cap = 500,
  weights_vulnerability = c(transmission = 0.50, outbreak = 0.30, susceptibility = 0.20),
  weights_priority = c(expected = 0.35, tail = 0.25, outbreak = 0.20, transmission = 0.20)
)

weight_sets <- score_params[c("weights_vulnerability", "weights_priority")]
for (weight_name in names(weight_sets)) {
  if (abs(sum(weight_sets[[weight_name]]) - 1) > 1e-9) {
    stop(paste("Weight set does not sum to 1:", weight_name))
  }
}

cap01 <- function(x, cap) {
  ifelse(is.na(x), 0, pmin(x / cap, 1))
}
```

### Assumptions and Limitations

- Caps are policy choices, not inferred epidemiologic constants.
- `vulnerability_score` and `priority_score` are decision-support summaries, not mechanistic parameters.
- `prob_large_outbreak` depends on the chosen threshold, currently `max(1% of susceptible population, 20 cases)`.
- Population-normalized burden metrics are stabilized with a population floor so tiny counties do not dominate the score simply because a few cases translate to extreme per-10k values.
- `tail_cases_per_10k` is still a Monte Carlo tail estimate and therefore noisier than means.

## 3. Simple Risk Score Formulation

For county $c$ in baseline scenario $r$, define capped components:

$$
\begin{aligned}
V_{c,r} &= \min\left(\frac{\max(R_v - 1, 0)}{c_{Rv}}, 1\right) \\
O_{c,r} &= P(\text{outbreak exceeds threshold}) \\
S_{c,r} &= \min\left(\frac{\text{susceptible \%}}{c_S}, 1\right) \\
E_{c,r} &= \min\left(\frac{\text{expected cases}}{c_E}, 1\right) \\
T_{c,r} &= \min\left(\frac{q_{95}(\text{cases})}{c_T}, 1\right)
\end{aligned}
$$

These are easier to interpret than percentile ranks and let the two scores answer different questions:

- $V$: how far above the herd-immunity threshold the county is
- $O$: how often an introduction grows into a large outbreak
- $S$: susceptible population share scaled to a chosen cap
- $E$: expected outbreak size in total cases
- $T$: high-end outbreak size in total cases

We then define two transparent score variants:

$$
\text{Vulnerability}_{c,r} = 0.50V_{c,r} + 0.30O_{c,r} + 0.20S_{c,r}
$$

$$
\text{Priority}_{c,r} = 0.35E_{c,r} + 0.25T_{c,r} + 0.20O_{c,r} + 0.20V_{c,r}
$$

```{r risk-formulation}
risk_input <- stochastic_results %>%
  left_join(
    simulation_stats_by_county %>%
      select(baseline_R0, county, p95_cases),
    by = c("baseline_R0", "county")
  ) %>%
  left_join(
    county_static_metrics %>%
      select(
        county,
        school_vax_rate,
        total_enrolled,
        total_vaccinated,
        immune_population,
        static_susceptible_population = susceptible_population,
        county_immunity_pct,
        county_susceptible_pct,
        R0_factor
      ),
    by = "county"
  ) %>%
  mutate(
    p95_cases = replace_na(p95_cases, 0),
    susceptible_population = coalesce(susceptible_population, static_susceptible_population),
    large_outbreak_threshold_cases = pmax(
      coalesce(
        large_outbreak_threshold_cases,
        susceptible_population * large_outbreak_threshold_frac
      ),
      large_outbreak_min_cases
    ),
    population_for_rate = pmax(population, score_params$per_10k_population_floor),
    expected_cases_per_10k = ifelse(population > 0, expected_cases / population * 10000, 0),
    tail_cases_per_10k = ifelse(population > 0, p95_cases / population * 10000, 0),
    expected_cases_per_10k_stabilized = expected_cases / population_for_rate * 10000,
    tail_cases_per_10k_stabilized = p95_cases / population_for_rate * 10000,
    rv_excess = pmax(vax_repnum - 1, 0),
    vulnerability_transmission_component = cap01(rv_excess, score_params$rv_excess_cap),
    vulnerability_outbreak_component = replace_na(prob_large_outbreak, 0),
    vulnerability_susceptibility_component = cap01(county_susceptible_pct, score_params$susceptible_pct_cap),
    priority_expected_component = cap01(expected_cases, score_params$expected_cases_cap),
    priority_tail_component = cap01(p95_cases, score_params$p95_cases_cap),
    priority_outbreak_component = replace_na(prob_large_outbreak, 0),
    priority_transmission_component = cap01(rv_excess, score_params$rv_excess_cap)
  )

risk_scores_internal <- risk_input %>%
  mutate(
    vulnerability_transmission_weighted =
      score_params$weights_vulnerability[["transmission"]] * vulnerability_transmission_component,
    vulnerability_outbreak_weighted =
      score_params$weights_vulnerability[["outbreak"]] * vulnerability_outbreak_component,
    vulnerability_susceptibility_weighted =
      score_params$weights_vulnerability[["susceptibility"]] * vulnerability_susceptibility_component,
    priority_expected_weighted =
      score_params$weights_priority[["expected"]] * priority_expected_component,
    priority_tail_weighted =
      score_params$weights_priority[["tail"]] * priority_tail_component,
    priority_outbreak_weighted =
      score_params$weights_priority[["outbreak"]] * priority_outbreak_component,
    priority_transmission_weighted =
      score_params$weights_priority[["transmission"]] * priority_transmission_component,
    vulnerability_score =
      vulnerability_transmission_weighted +
      vulnerability_outbreak_weighted +
      vulnerability_susceptibility_weighted,
    priority_score =
      priority_expected_weighted +
      priority_tail_weighted +
      priority_outbreak_weighted +
      priority_transmission_weighted
  ) %>%
  group_by(baseline_R0) %>%
  mutate(
    vulnerability_score_center = mean(vulnerability_score, na.rm = TRUE),
    vulnerability_score_scale = sd(vulnerability_score, na.rm = TRUE),
    vulnerability_relative_mean1 = ifelse(vulnerability_score_center > 0, vulnerability_score / vulnerability_score_center, 1),
    vulnerability_relative_z = ifelse(vulnerability_score_scale > 0, (vulnerability_score - vulnerability_score_center) / vulnerability_score_scale, 0),
    priority_score_center = mean(priority_score, na.rm = TRUE),
    priority_score_scale = sd(priority_score, na.rm = TRUE),
    priority_relative_mean1 = ifelse(priority_score_center > 0, priority_score / priority_score_center, 1),
    priority_relative_z = ifelse(priority_score_scale > 0, (priority_score - priority_score_center) / priority_score_scale, 0)
  ) %>%
  ungroup()
```

## 4. Internal Per-County Breakdown (Restricted)

This table now shows raw metrics, capped score inputs, and the two primary score variants.

```{r county-breakdown-internal}
risk_county_breakdown_baseline <- risk_scores_internal %>%
  filter(baseline_R0 == R0_baseline) %>%
  arrange(desc(vulnerability_score), desc(priority_score)) %>%
  transmute(
    county,
    population,
    school_vax_rate,
    county_immunity_pct,
    county_susceptible_pct,
    susceptible_population,
    large_outbreak_threshold_cases,
    vax_repnum,
    rv_excess,
    prob_large_outbreak,
    expected_cases,
    expected_cases_per_10k,
    expected_cases_per_10k_stabilized,
    p95_cases,
    tail_cases_per_10k,
    tail_cases_per_10k_stabilized,
    vulnerability_transmission_component,
    vulnerability_outbreak_component,
    vulnerability_susceptibility_component,
    priority_expected_component,
    priority_tail_component,
    priority_outbreak_component,
    priority_transmission_component,
    vulnerability_score,
    priority_score,
    vulnerability_relative_mean1,
    vulnerability_relative_z,
    priority_relative_mean1,
    priority_relative_z
  )

knitr::kable(
  risk_county_breakdown_baseline %>%
    mutate(
      population = comma(population),
      school_vax_rate = round(school_vax_rate, 3),
      county_immunity_pct = round(county_immunity_pct, 3),
      county_susceptible_pct = round(county_susceptible_pct, 3),
      susceptible_population = comma(susceptible_population),
      large_outbreak_threshold_cases = round(large_outbreak_threshold_cases, 1),
      vax_repnum = round(vax_repnum, 3),
      rv_excess = round(rv_excess, 3),
      prob_large_outbreak = round(prob_large_outbreak, 4),
      expected_cases = round(expected_cases, 2),
      expected_cases_per_10k = round(expected_cases_per_10k, 3),
      expected_cases_per_10k_stabilized = round(expected_cases_per_10k_stabilized, 3),
      p95_cases = round(p95_cases, 1),
      tail_cases_per_10k = round(tail_cases_per_10k, 3),
      tail_cases_per_10k_stabilized = round(tail_cases_per_10k_stabilized, 3),
      vulnerability_transmission_component = round(vulnerability_transmission_component, 3),
      vulnerability_outbreak_component = round(vulnerability_outbreak_component, 3),
      vulnerability_susceptibility_component = round(vulnerability_susceptibility_component, 3),
      priority_expected_component = round(priority_expected_component, 3),
      priority_tail_component = round(priority_tail_component, 3),
      priority_outbreak_component = round(priority_outbreak_component, 3),
      priority_transmission_component = round(priority_transmission_component, 3),
      vulnerability_score = round(vulnerability_score, 3),
      priority_score = round(priority_score, 3),
      vulnerability_relative_mean1 = round(vulnerability_relative_mean1, 3),
      vulnerability_relative_z = round(vulnerability_relative_z, 3),
      priority_relative_mean1 = round(priority_relative_mean1, 3),
      priority_relative_z = round(priority_relative_z, 3)
    ),
  caption = paste0("Internal county-level metric breakdown for baseline R0 = ", R0_baseline)
)
```

## 5. Risk Score Definitions

The model now reports two primary score variants:

- `vulnerability_score`: emphasizes the conditions for sustained transmission and is intentionally close to population-neutral.
- `priority_score`: emphasizes expected and upper-tail burden, while still preserving some weight on outbreak probability and transmission potential.
  It is the more operational ranking.

For communication, the raw metrics should usually be shown first and the two scores second.

## 6. County Risk Scores Only

```{r county-scores-only}
risk_scores_only_baseline <- risk_scores_internal %>%
  filter(baseline_R0 == R0_baseline) %>%
  arrange(desc(vulnerability_score), desc(priority_score)) %>%
  transmute(
    county,
    vulnerability_score,
    priority_score,
    vulnerability_relative_mean1,
    vulnerability_relative_z,
    priority_relative_mean1,
    priority_relative_z
  )

knitr::kable(
  risk_scores_only_baseline %>%
    mutate(
      vulnerability_score = round(vulnerability_score, 3),
      priority_score = round(priority_score, 3),
      vulnerability_relative_mean1 = round(vulnerability_relative_mean1, 3),
      vulnerability_relative_z = round(vulnerability_relative_z, 3),
      priority_relative_mean1 = round(priority_relative_mean1, 3),
      priority_relative_z = round(priority_relative_z, 3)
    ),
  caption = paste0("County score variants for baseline R0 = ", R0_baseline)
)
```

## 7. Scores Across R0 Scenarios

```{r risk-across-r0-internal}
risk_score_summary_by_r0 <- risk_scores_internal %>%
  group_by(baseline_R0) %>%
  summarise(
    mean_vulnerability = mean(vulnerability_score, na.rm = TRUE),
    p25_vulnerability = quantile(vulnerability_score, 0.25, na.rm = TRUE),
    p50_vulnerability = quantile(vulnerability_score, 0.50, na.rm = TRUE),
    p75_vulnerability = quantile(vulnerability_score, 0.75, na.rm = TRUE),
    mean_priority = mean(priority_score, na.rm = TRUE),
    p25_priority = quantile(priority_score, 0.25, na.rm = TRUE),
    p50_priority = quantile(priority_score, 0.50, na.rm = TRUE),
    p75_priority = quantile(priority_score, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

score_summary_long <- bind_rows(
  risk_score_summary_by_r0 %>%
    transmute(
      baseline_R0,
      score_type = "Vulnerability",
      mean_score = mean_vulnerability,
      p25_score = p25_vulnerability,
      p50_score = p50_vulnerability,
      p75_score = p75_vulnerability
    ),
  risk_score_summary_by_r0 %>%
    transmute(
      baseline_R0,
      score_type = "Priority",
      mean_score = mean_priority,
      p25_score = p25_priority,
      p50_score = p50_priority,
      p75_score = p75_priority
    )
)

ggplot(score_summary_long, aes(x = baseline_R0, y = p50_score, color = score_type, fill = score_type)) +
  geom_ribbon(aes(ymin = p25_score, ymax = p75_score), alpha = 0.20, color = NA) +
  geom_line(linewidth = 1.0) +
  geom_line(aes(y = mean_score), linewidth = 0.8, linetype = "dashed") +
  labs(
    title = "Vulnerability and Priority Scores Across Baseline R0",
    x = "Baseline R0",
    y = "Score"
  ) +
  theme_minimal()
```

## 8. Frontend-Ready Metrics

This flat file is intended to be passed directly to a frontend. It keeps the raw
county metrics visible rather than forcing the UI to depend on a single composite
score. Static county metrics are repeated across baseline $R_0$ scenarios so the
frontend can filter and display from one CSV.

```{r frontend-risk-metrics}
frontend_risk_metrics <- risk_scores_internal %>%
  transmute(
    baseline_R0,
    county,
    population,
    susceptible_population,
    large_outbreak_threshold_cases,
    total_enrolled,
    total_vaccinated,
    school_vax_rate,
    immune_population,
    county_immunity_pct,
    county_susceptible_pct,
    R0_factor,
    vax_repnum,
    rv_excess,
    expected_cases,
    expected_cases_per_10k,
    expected_cases_per_10k_stabilized,
    prob_large_outbreak,
    p95_cases,
    tail_cases_per_10k,
    tail_cases_per_10k_stabilized,
    vulnerability_transmission_component,
    vulnerability_outbreak_component,
    vulnerability_susceptibility_component,
    priority_expected_component,
    priority_tail_component,
    priority_outbreak_component,
    priority_transmission_component,
    vulnerability_score,
    priority_score,
    vulnerability_relative_mean1,
    vulnerability_relative_z,
    priority_relative_mean1,
    priority_relative_z
  ) %>%
  arrange(baseline_R0, desc(vulnerability_score), desc(priority_score), county)

frontend_risk_metrics_baseline <- frontend_risk_metrics %>%
  filter(baseline_R0 == R0_baseline)
```

## 9. Export Outputs

```{r export-risk-scores, cache=FALSE}
export_dir <- file.path("outputs", "utah_county_risk")
if (!dir.exists(export_dir)) dir.create(export_dir, recursive = TRUE)

# Internal all-scenario scores
write.csv(
  risk_scores_internal,
  file.path(export_dir, "risk_scores_internal_all_r0.csv"),
  row.names = FALSE
)

# Internal county-level breakdown for baseline scenario
write.csv(
  risk_county_breakdown_baseline,
  file.path(export_dir, "risk_scores_internal_county_breakdown_baseline_r0.csv"),
  row.names = FALSE
)

write.csv(
  risk_scores_only_baseline,
  file.path(export_dir, "risk_scores_internal_only_baseline_r0.csv"),
  row.names = FALSE
)

write.csv(
  risk_score_summary_by_r0,
  file.path(export_dir, "risk_score_summary_by_r0.csv"),
  row.names = FALSE
)

write.csv(
  frontend_risk_metrics,
  file.path(export_dir, "frontend_risk_metrics_all_r0.csv"),
  row.names = FALSE
)

write.csv(
  frontend_risk_metrics_baseline,
  file.path(export_dir, "frontend_risk_metrics_baseline_r0.csv"),
  row.names = FALSE
)

# Internal (contains county names): keep private and do not publish
save(
  risk_scores_internal,
  risk_county_breakdown_baseline,
  risk_scores_only_baseline,
  risk_score_summary_by_r0,
  frontend_risk_metrics,
  frontend_risk_metrics_baseline,
  score_params,
  file = file.path(export_dir, "risk_scores_internal_private.RData")
)

cat("Saved risk score outputs to:", normalizePath(export_dir), "\n")
cat("Frontend-ready metric CSVs now include school vaccination, immunity, susceptibility, and simulation outputs.\n")
```
